### 2020-10-13 latex/manuscript.tex

\documentclass{article}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks

% tables
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{multirow}

% floats
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}

% math
\usepackage{amsfonts,amsmath,amssymb}
\usepackage{nicefrac}

\renewcommand{\Pr}{\operatorname{Pr}}
\newcommand{\E}{\ensuremath{\operatorname{E}}}
\newcommand{\V}{\operatorname{V}}
\newcommand{\st}{\operatorname{subject\ to:\ }}
\newcommand{\argmin}[1]{\operatornamewithlimits{argmin:\ }_{#1}}
\newcommand{\argmax}[1]{\operatornamewithlimits{argmax:\ }_{#1}}

\newcommand{\CP}{\ensuremath{\operatorname{CP}}}
\newcommand{\ACP}{\ensuremath{\operatorname{ACP}}}
\newcommand{\OCP}{\ensuremath{\operatorname{OCP}}}
\newcommand{\PP}{\ensuremath{\operatorname{PP}}}
\newcommand{\EP}{\ensuremath{\operatorname{EP}}}
\renewcommand{\Pr}{\ensuremath{\operatorname{Pr}}}
\newcommand{\cond}{\ensuremath{\,|\,}}
%\newcommand{\c}{\ensuremath{z_{1-\alpha}}}

% other
\usepackage{color}
\usepackage{natbib}
\usepackage{scalerel}
\newcommand{\kr}[1]{{\color{red}#1}}

\title{Conditional Power and Friends: The Why and How of (Un)planned, Unblinded Sample Size Recalculations in Confirmatory Trials}

\author{
  Kevin Kunzmann \thanks{
        MRC Biostatistics Unit, University of Cambridge,
        Cambridge Institute of Public Health,
        Forvie Site, Robinson Way,
        Cambridge Biomedical Campus,
        Cambridge CB2 0SR,
        United Kingdom
    }\\
    MRC Biostatistics Unit \\
    University of Cambridge \\
    \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk} \\

    Michael J. Grayling \\
    Population Health Sciences Institute \\
    Newcastle University \\
    \texttt{michael.grayling@newcastle.ac.uk} \\

    Kim May Lee \\
    Pragmatic Clinical Trials Unit\\
    Queen Mary University of London\\
    \texttt{k.m.lee@qmul.ac.uk} \\

    David S.\ Robertson \\
    MRC Biostatisics Unit\\
    University of Cambridge\\
    \texttt{david.robertson@mrc-bsu.cam.ac.uk} \\

    Kaspar Rufibach \\
    Methods, Collaboration, and Outreach Group (MCO) \\
    Department of Biostatistics\\
    F. Hoffmann-La Roche, Basel\\
    \texttt{kaspar.rufibach@roche.com} \\

    James M. S. Wason \\
    Population Health Sciences Institute \\
    Newcastle University\\
    and \\
     MRC Biostatistics Unit \\
    University of Cambridge \\
    \texttt{james.wason@newcastle.ac.uk}
}

\usepackage{setspace}
\onehalfspacing
\pdfminorversion=4

\begin{document}

\maketitle

\captionsetup{width=\textwidth}

%\newpage
%\section*{Funding}
%\noindent
%David S. Robertson was funded by the Biometrika Trust and the Medical Research %Council under Grant MC\_UU\_00002/6.

\newpage

\begin{abstract}
Determining the sample size of a confirmatory trial is challenging.
If the sample size is too small, the chance of detecting an effect is small.
If it is too big, trial duration and costs are needlessly inflated.
Adapting the final sample size of a trial to the evidence accruing 
during the trial is thus a natural way to address 
this planning uncertainty.
In a regulatory environment, 
designs with adaptive sample size need to 
account for their optional stopping to guarantee
strict type-I error-rate control. 
A variety of different methods to maintain type-I error-rate control
after unplanned changes of the initial sample size have been proposed 
in the literature rendering interim analyses for
the purpose of sample size recalculation feasible in
practice.
Any form of interim analysis allows the analysis of
a trial's prospects to finally reject the null hypothesis 
given the data observed up to the interim analysis.
Since the sample size is usually determined via an 
argument based on the power of the trial, 
an interim analysis raises the question of
how the final sample size should be determined conditional on the accrued information.
Conditional power is a concept often put forward in this context.
Since it depends on the unknown effect size, 
we take a strict estimation perspective and compare
assumed conditional power, observed conditional power, 
and predictive power with respect to their properties as estimators 
of the unknown conditional power.
We then demonstrate that pre-planning an interim analysis using 
methodology for unplanned interim analyses is ineffective and 
naturally leads to the concept of optimal two-stage designs.
We conclude that unplanned design adaptations should only be conducted
as reaction to trial-external new evidence, operational needs to violate the originally chosen design, 
or \textit{post~hoc} changes in the objective criterion.
Finally, we show that commonly discussed sample size recalculation rules can lead to paradoxical outcomes and propose two alternative ways of reacting to
newly emerging trial-external evidence.
\end{abstract}

Keywords: Adaptive Design - Conditional Power - Interim Analysis - Optimal Design - Predictive Power - Sample Size Recalculation

\newpage

\section{Introduction}
\label{sec:introduction}

The planning phase of confirmatory clinical trials is typically 
characterized by substantial uncertainty about the magnitude of the 
parameters underlying the hypothesis of interest.
Often the alternative hypothesis is the superiority of a new intervention arm
over a treatment-as-usual control arm and is formalized via
the average treatment effect, e.g., the difference of means between the 
two randomly allocated treatment arms. 
Clinical data collection is expensive and time consuming leading to a
strong economic incentive to reach the study goals with as little data as possible.
The conventional statistical criteria to determine the sample size of a trial 
are a one-sided type-I error rate of
2.5\% and a power of 80\% or 90\%.
Since power is a function of the unknown effect size the initial design 
must be specified under substantial uncertainty about the magnitude of the effect which the trial hopes to detect eventually.
Mainly, two ways of addressing this challenge have been put forward in the literature.

Firstly, the so-called `hybrid' approach to sample size 
derivation takes a Bayesian
view on determining the initial sample size of a clinical trial~\citep{spiegelhalter1994}.
The Bayesian component requires the specification of an informative 
prior on the parameters of interest.
This then allows reasoning about the `expected power' of a trial as a
function of its sample size and, consequently, determination of the sample size
such that the expected power exceeds a target threshold. 
This concept is a straight-forward extension of the usual practice of 
computing the power under a fixed point alternative, which can easily be recovered as a special case when considering a point prior.
The advantage lies in the fact that the \textit{a~priori} information about the magnitude of the effect size is faithfully 
reflected in the sample size derivation.
Since the actual analysis is still frequentist, type-I error rate control is not compromised.

Secondly, authors have proposed to apply the concept of adaptive design changes
to recalculate the initial sample size of a design based on data observed 
within the trial itself, see e.g.~\citet{bauer2016}.
The rationale behind this approach is to use the accruing evidence about the unknown parameters driving the sample size derivation to 
`correct' the sample size mid-trial. 
During the interim analysis, the accrued data can either be unblinded
or not.
Methods for the latter `blinded interim analysis' are particularly useful if relevant nuisance parameters such as the variance are also unknown \citep{birkett1994internal,bauer2016}.
We focus on the unblinded case which allows for a more precise 
interim assessment of the effect size than methods that retain 
the blind. 
The maximal type-I error rate constraint is then usually protected by 
applying the conditional error principle~\citep{muller2004,brannath2012}
or an equivalent formulation via $\alpha$-spending or 
combination functions~\citep{bauer2016}.
Often the sample size of the current 
trial is adjusted such that the conditional power given the data observed up to the
interim analysis again exceeds the initial threshold for unconditional 
power~\citep{proschan1995}.
Any recalculation based on conditional power arguments must
address the problem that conditional power, 
just as unconditional power, 
depends on the unknown underlying parameters.
It must thus be
estimated to inform a sample size recalculation.
Yet, precise estimation of the relevant parameters before the conclusion of a trial is hard since only a fraction of the final sample size is available.
Three approaches to estimating conditional power have been discussed in the literature.

Firstly, 
`conditional power' is the probability to reject the null hypothesis given the interim data as a function of the unknown parameters.
In a slight abuse of terminology, the same term is often used to refer to
the assumed conditional power that is obtained by plugging in the 
point alternative used for the initial sample size derivation. 
Evidently, this assumed conditional makes no use of the accrued trial
data since the effect size is kept fixed.
Secondly,
authors have proposed to use the `observed conditional power' instead, which replaces the unknown parameters with their maximum-likelihood estimates.
The latter `plug-in' approach is often criticized for 
`... [using] the interim estimate of the effect [...] twice ...' \citep[p.~330]{bauer2016} (see also~\cite{bauer2006}).
Thirdly, 
conditional power can be evaluated as Bayesian expected power conditional on the observed interim data, 
i.e., by averaging conditional power as function of the unknown 
parameters with respect to the posterior density 
after conducting the interim analysis. 
Within the hybrid Bayesian framework this is usually referred to as 
`predictive power'~\citep{spiegelhalter1994,bauer2016}.

The purpose of this paper is to argue that pre-planned sample size adaptations
using methodology intended for unplanned interim analyses to react to within-trial interim data are inefficient and unnecessary if the original design was planned optimally. 
We then discuss cases where an unplanned design adaptation might still be warranted and propose two criteria that allow such an adaptation that is
consistent with the original design.
We first review assumed \mbox{conditional-,} observed \mbox{conditional-,} and predictive power 
from an estimation perspective.
For sake of simplicity, 
we use a simple single-arm one-stage design to illustrate core 
characteristics.
We then discuss the risks of constructing an 
`adaptive' two-stage design by na\"{\i}vely applying
a conditional-power-based sample size recalculation rule and the
conditional error principle.
The drawbacks of this na\"{\i}ve approach directly lead to the
concept of optimal-two stage designs and 
we build on ideas presented in \cite{brannath2004,pilz2019}, and
\cite{adoptrjss2020} to derive optimal designs for the situation studied in
this manuscript.
Without loss of generality, we focus on minimal expected sample size as optimality criterion.
Optimal two-stage designs, by definition, cannot be made more
efficient by sample size recalculation.
The optimality of such designs does, however, depend on the initial trial-external evidence that feeds into the planning assumptions.
This trial-external evidence might change during an ongoing study 
and may thus mandate a sample size recalculation.
In Section~\ref{sec:unplanned-adaptation}, 
we discuss when an unplanned sample size recalculation is 
reasonable, 
how such a recalculation interacts with the concept of 
optimal two-stage designs,
and the practical implementation.

In the following, 
we assume that the interest lies in testing a new treatment for efficacy.
For the sake of simplicity,
we consider the case of a single-arm trial.
All considerations can easily be extended to the more 
practically relevant two-arm case. 
We further assume that the individually observed outcomes of the study participants $X_i, i = 1,\ldots,n$ are $iid$ and that their distribution has finite first moment $\theta$ and unit variance.
Again, all considerations can be extended to the case of 
generic known variances and, 
at least approximately, to the case of unknown variance. 
A suitable test statistic for the null hypothesis of interest~${\mathcal{H}_0:\theta\leq0}$ is
\begin{align}
    Z_n = \frac{1}{\sqrt{n}} \sum_{i=1}^n X_i \ .
\end{align}
Invoking the central limit theorem,
$Z_n\stackrel{\cdot}{\sim}\mathcal{N}(\sqrt{n}\,\theta, 1)$
and, 
on the boundary of the null hypothesis, $Z_n\stackrel{\cdot}{\sim}\mathcal{N}(0, 1)$.

Further assuming a maximal permissible type-I error rate of $\alpha=0.025$ (which we use for the remainder of the paper), the critical value for a single-stage fixed-sample-size design is the
$1-\alpha$ quantile of the standard normal distribution, i.e., approximately $c=1.96$.
The test then rejects $\mathcal{H}_0$ if and only if $Z_n > c$ after the outcomes of $n$ subjects have been observed. 
The required size of the trial $n$, given a maximal allowable type-I error
rate, is usually determined by some form of restriction on 
the (minimal) statistical power of the test.
Approaches to defining such a power constraint under 
different assumptions were reviewed in~\citet{kunzmann2020}.

\section{Monitoring power} %%%%%%%%%%%%%%%%%%%%%%%
\label{sec:monitoring-power}

No matter which rationale was used to derive the overall sample size $n$,
after observing $m$, $0<m<n$, outcomes,
an independent data monitoring committee might be interested to learn about the prospects of
eventually rejecting the null hypothesis.
Both the rejection probability conditional on the presence of an effect (conditional power) or the joint probability of rejection and presence of an effect (conditional probability of success) could be of interest.
Here, we focus on the former due to the central role of
power (and conditional power) in the planning of clinical trials.
This effectively constitutes an estimation problem where the estimand is 
the \emph{conditional} probability of rejecting the null hypothesis given
$Z_m=z_m$ and $\theta > 0$.
Here the conditioning on a positive effect is crucial to justify the notion
of `power' in contrast to the unconditional `probability of success'~\citep{kunzmann2020}.

Since $Z_n$ and $Z_m$ are (asymptotically) jointly normal, the conditional distribution of $Z_n$ given $Z_m$ and $\theta$ is again normal and 
given by
\begin{align}
    \mathcal{L}_\theta\big[\, Z_n \mid Z_m = z_m \,\big] \stackrel{\cdot}{=} \mathcal{N}\Big(\sqrt{n}\,\theta + \sqrt{\tau}\big(z_{m} - \sqrt{m}\,\theta\big),\ 1 - \tau\Big) \label{eq:joint-distribution}
\end{align}
where $\tau:=m/n$ is the `information fraction' at the interim analysis.

The probability of rejecting the null hypothesis at the end of the trial 
as a function of $\theta$ is referred to as the `conditional power' in the literature.
It is defined as
\begin{align}
    \CP(z_m, c, \theta) :&= \Pr_\theta[\,Z_n > c \cond Z_m = z_m\,] \\
        &= 1 - \Phi\left(
            \frac{c - \sqrt{n}\,\theta - \sqrt{\tau} \, z_m + \sqrt{\tau}\sqrt{m}\,\theta
            }{\sqrt{1 - \tau}}
        \right) \\
        &= 1 - \Phi\left(
            \frac{c/\sqrt{m} + \sqrt{\tau}\,\big(\theta - \widehat{\theta}_m\big) - \theta/\sqrt{\tau}
            }{\sqrt{(n - m)/(n\,m)}}
        \right)      \ .
\end{align}
Ignoring the implicit dependency on the time point of the interim analysis, $m$, 
conditional power is thus a function of the observed effect $\widehat{\theta}_m:=z_m/\sqrt{m}$ 
(or equivalently the observed test statistic $z_m$), the critical value for the test decision~$c$, 
and the standardized effect size~$\theta$.  
Since $\theta$ is unknown, 
so is $\CP(z_m, c, \theta)$ and it thus cannot be 
evaluated directly upon observing $Z_m=z_m$.
Yet, as a function of the unknown quantity $\theta$, 
$\CP(z_m, c, \theta)$ can be estimated from observed data.
The estimation perspective on `evaluating' conditional power is 
less commonly taken in the literature but provides a consistent framework to
compare the characteristics of different methods 
(see also~\citep[Discussion]{bauer2006}).

Firstly, conditional power can be estimated based on a fixed point alternative~$\theta_1$.
In a slight abuse of terminology, 
this quantity is often also referred to as `conditional power'.
To clearly distinguish it from $\CP(z_m, c, \theta)$ we denote 
it `assumed conditional power'
\begin{align}
    \ACP(z_m,c) := \CP\big(\, z_m, c, \theta_1 \,\big) \ .
\end{align}

Secondly, the observed effect $\widehat{\theta}_m$ can be used 
as a plug-in estimator for the unknown effect size
(see, e.g.~\citep{proschan1995}).
This quantity is sometimes referred to as `observed conditional power'
in the literature and is defined as
\begin{align}
    \OCP(z_m, c) := \CP\big(\, z_m, c, \widehat{\theta}_m \,\big) \ .
\end{align}

Thirdly, a Bayesian approach can be taken if one 
is willing to quantify the uncertainty about the unknown parameter $\theta$ by modeling it as the
realization of a random variable $\Theta\sim\varphi(\cdot)$ where $\varphi(\theta)$ is the
prior probability density function evaluated at the parameter value $\theta$.
Our definition of this so-called `predictive power'
\begin{align}
    \PP_\varphi(z_m,c) :&= \Pr_\varphi[\,Z_n > c \cond \Theta >0, Z_m = z_m\,] \\ 
    &= \int_{0}^{\infty}\CP\big(z_m, c, \theta\big)\,\varphi\big(\theta\cond Z_m=z_m, \Theta\geq0\big)\operatorname{d}\theta
\end{align}
differs slightly from the one proposed by~\citet{spiegelhalter1994}
in that we condition on a positive effect size $\Theta>0$. 
This is more consistent with the notion of frequentist power as discussed in~\cite{kunzmann2020}, although the difference is 
only of practical relevance when a substantial
fraction of the \textit{a~priori} probability mass is concentrated on the null hypothesis.

We now compare $\ACP, \OCP$, and $\PP$ by means of a concrete example.
To this end we assume that the available prior information can be summarized in a truncated normal prior with density
\begin{align}
    \varphi(\theta) := \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\frac{\displaystyle \phi\left(\frac{\theta - 0.4}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - 0.4}{0.2}\right) - \Phi\left(\frac{-0.5 - 0.4}{0.2}\right)} \ ,
\end{align}
(see Figure~\ref{fig:prior-pdf})
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/prior_density}
    \caption{%
        Assumed prior density function. 
        The gray area indicates the null hypothesis, $\mathcal{H}_0$, of no effect. 
    }
    \label{fig:prior-pdf}
\end{figure}
and that all positive effect sizes are clinically relevant.
A truncated normal prior is convenient since it allows the analytic
computation of the posterior distribution under a normal likelihood.
It is also the maximum entropy distribution on a compact interval given mean and standard deviation and thus `least informative'
given a lower and upper boundaries on plausible effects as well as
the location and vagueness (variance) of the prior.
Following~\citep{kunzmann2020} the required 
sample size $n$ is then determined by requiring 
a minimal expected power of $1-\beta=80\%$, 
where expected power is defined as 
\begin{align}
    \EP_\varphi(n, c) :=
    \Pr\big[\,Z_n>c\,|\,\Theta\geq0\,\big] = \int \Pr_\theta\big[\,Z_n>c\,\big]\, \varphi(\theta\,|\,\Theta\geq0) \,\operatorname{d}\theta \ .
\end{align}
Under these assumptions, the required sample size is 
$n = 79$.
No point alternative is necessary to derive this sample size.
We thus use the prior mean rounded to the first decimal point to 
evaluate $\ACP$ and define $\theta_1:=0.4$.

The three estimators are depicted in Figure~\ref{fig:acp-ocp-pp-comparison-absolute} as functions 
of the observed interim outcome, where the interim time-point $m = 26$.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_example}
        \caption{%
            Estimates as function of the interim data.
        }
        \label{fig:acp-ocp-pp-comparison-absolute}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/acp_ocp_pp_sampling_distributions}
        \caption{%
            Histograms of the sampling distributions.
        }
        \label{fig:sampling-distributions}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/acp_ocp_pp_bias_mae_mse}
        \caption{%
            Bias, mean absolute error (MAE) and mean squared error (MSE).
        }
        \label{fig:acp-ocp-pp-as-estimators}
    \end{subfigure}
    \caption{%
        Properties of $\ACP,\OCP$, and $\PP$ as estimators of the unknown conditional power at $m=26$ an overall sample size $n=79$.
    }
\end{figure}
The most sensitive measure is $\OCP$ while $\ACP$ is the least sensitive
(see also spread of the sampling distributions in Figure~\ref{fig:sampling-distributions}). 
This is due to $\ACP$ assuming a fixed effect size and thus 
only being affected by direct changes in $z_m$.
Both $\OCP$ and $\PP$, however, are also indirectly affected by updating 
the belief about the effect size with the interim results.
The plug-in estimate $\OCP$ assumes that the observed effect is the 
true effect and does not quantify the uncertainty around this value in any way. 
$\PP$, on the other hand, invokes Bayes' theorem and then integrates 
$\CP(z_m,c,\theta)$ with respect to the obtained posterior distribution. 
The degree of adaptation of $\PP$ thus depends 
on the vagueness of the chosen prior. 
In fact, as the prior approaches a point mass at $\theta_1$, $\PP$ converges
to $\ACP$.
$\ACP$ can thus be seen as a mere special case of $\PP$ with a point prior
on $\theta_1>0$.

A fundamental difference between $\ACP$ and $\PP$ on the one hand and $\OCP$ on the other hand is that
$\OCP$ is the only estimator that does not condition on~$\Theta$.
In contrast, $\ACP$ implicitly conditions on $\Theta=\theta_1>0$ and 
$\PP$~on~${\Theta > 0}$.
The practical implication of this can be seen 
by looking at the sampling distributions of the 
three estimators under different assumptions 
about the value of $\theta$ (see~Figure~\ref{fig:sampling-distributions}).
Since $\OCP$ does not condition on $\Theta>0$, the sampling distribution is much more left-skewed for small effects than for the other two estimators.
Also, the negligence of the sampling variation of $\widehat{\theta}_m$, which $\OCP$ plugs in the expression for conditional power, leads to a larger variance for intermediate values of $\theta$ and the characteristic
U-shape \citep{bauer2006}.
This high variance of $\OCP$ directly translates to a relatively large mean absolute error (MAE) and mean squared error (MSE) when estimating conditional power, 
see Figure~\ref{fig:acp-ocp-pp-as-estimators}.
$\PP$ is the posterior expectation of conditional power with respect to the chosen prior and thus  minimizes the quadratic Bayes risk, i.e.~the MSE.
This leads to relative good precision for parameter values with high \textit{a~priori} likelihood 
(around $\theta=0.4$).
If one wanted to minimize the MAE directly, 
the posterior median would be optimal.
The principle, however, remains the same and
the posterior mean is more consistent with the derivation of the initial sample size using expected power. 
$\ACP$ is clearly the best in terms of bias and MAE/MSE for values close to or
above $\theta_1$, but its performance as estimator of the conditional power 
quickly deteriorates for small effect sizes.

$\ACP$ is merely an extreme special case of $\PP$. 
$\OCP$ is both hard to justify theoretically and has inferior precision
for parameter values with a high \textit{a~priori} likelihood.
$\PP$ is thus the natural choice for monitoring power during the course of a trial.
Depending on the prior chose, it's properties can be either more similar to $\ACP$ (low prior variance) or $\OCP$ (high prior variance).

\section{Na\"{\i}ve unplanned sample size adaptations}
\label{sec:naive-adaptations}

Monitoring the predictive power of an ongoing trial naturally raises the question of whether this information can be leveraged to improve the operating characteristics of a trial.
Typically, sample size recalculation is considered in the context of group-sequential designs at the pre-specified interim analysis \citep{proschan1995,brannath2004,bauer2016}.
From a statistical perspective, 
there is no reason why a sample size recalculation should not be
performed for a study that initially was planned using a single-stage design.
For the sake of simplicity the following considerations thus assume a simple 
single-stage design with fixed $n$ and $c$ as starting point and are
restricted to a single interim analysis.
All results can be generalized to the case of more complex starting designs and multiple interim analyses.

A method for sample size recalculation that is often discussed in the literature is to derive a new sample size $n'$ and a new critical value $c'$ such that some estimate of conditional power exceeds a threshold $1-\beta_{cond.}$.
Different choices for the estimator of conditional power
and the choice of $\beta_{cond.}$ have been discussed in the literature \citep{proschan1995,brannath2004,bauer2016}.
Strict overall type-I error rate control can be maintained by invoking the conditional error principle \citep{muller2001,muller2004,brannath2012}, i.e., by limiting the maximal type-I error rate of the new design to the maximal conditional type-I error rate of the original design given the observed $z_m$.
Often, trial protocols leave the exact choice of $1-\beta_{cond.}$ open since it can be chosen \textit{ad~hoc}
without compromising strict type-I-error rate control.
The basic concept of readjusting the sample size to achieve the desired conditional power is, however, a common approach, see for instance,~\citet{bhatt2013} and~\citet{mehta2009}.

In accordance with the discussion in the previous section
and to be consistent with the derivation of the
initial sample size of $n=79$ in the example considered earlier, 
we use predictive power as an estimator of conditional power.
Furthermore, we set $\beta_{cond.}=\beta=0.2$.
For given $Z_m = z_m$, the recalculation rule then corresponds to solving 
the optimization problem
\begin{align}
    \argmin{n', c'} &&                   n' & \label{eq:naive-recalc}\\
    \st             && \CP(m, n', z_m, c', 0) &\ \leq \ \CP_n(m, n, z_m, c, 0) \label{eq:ce-principle-naive-recalc} \\
                    &&    \PP_\varphi(m, n', z_m, c') &\ \geq \ 1 - \beta_{cond.} \\
                    &&                   n' &\ \geq \ n_{min} > m \label{eq:nmin-naive-recalc} \\
                    &&                   n' &\ \leq \ n_{max} \label{eq:nmax-naive-recalc} \ . 
\end{align}
Here, we indicate the dependency on the 
final sample size and the interim time point explicitly by redefining
\begin{align}
    \CP(m, n, z_m, c, \theta) :&= \Pr_\theta[\,Z_n > c \cond Z_m = z_m\,] \\
    \PP_\varphi(m, n, z_m, c) :&= \Pr_\varphi[\,Z_n > c \cond \Theta >0, Z_m = z_m\,] \ .
\end{align}
Constraint~\eqref{eq:ce-principle-naive-recalc} implements the conditional
error principle by limiting the conditional (type-I) error rate under the new design to 
the conditional (type-I) error rate under the original design.
The minimal sample size constraint~\eqref{eq:nmin-naive-recalc} allows  $n_{min}>m$ for cases where a minimal sample size upon rejection of the null hypothesis is deemed necessary.
This could e.g.~be of interest if a minimal precision of the final maximum likelihood estimator is sought.
Note that the trial cannot stop at $n'=m$ without immediately accepting the
null hypothesis since the conditional error under the new design in the case of early acceptance would be $1$ but is always smaller than $1$ for the original single-stage design.
Imposing a maximal sample size of $n_{max}$ via constraint~\eqref{eq:nmax-naive-recalc} is a practical necessity since the 
recalculated sample size could otherwise tend to infinity as $z_m$ approaches negative infinity.
In cases where this constraint prevents a solution (low observed effect), 
the trial is usually stopped early for futility declaring that the null hypothesis cannot be rejected. 
Alternatively, one could continue with the maximal sample size accepting
the fact that the conditional power constraint is not met.
This would, however, raise the question as to why a hard constraint on conditional power was imposed in the first place.

If the recalculation rule defined in \eqref{eq:naive-recalc}--\eqref{eq:nmax-naive-recalc} is made mandatory in the
study protocol, 
it yields two functions $n(z_m)$ and $c(z_m)$ as the point-wise solution of the problem
which jointly define an `adaptive design' for any $z_m$.
Here, `mandatory' means that it is decided \textit{a~priori} to always (for all $z_m$)
recalculate the 
final sample size and critical value in the 
above specified way after a fixed number of outcomes $m$ has been observed.
The term `adaptive design' is slightly misleading in this context.
The design itself is pre-specified (and thus not adapted or changed) 
but only $n$ and $c$ are adaptive since they vary as functions of $z_m$.
We would like to emphasize that such a detailed specification of the sample size recalculation rule is not common in practice since most protocols are focused on guaranteeing control of the maximal type-I-error rate control and maintaining maximal flexibility during the interim analysis.
Still, if a sample-size reassessment was prescribed in the study protocol, there must have been some rationale as to the goal of that reassessment and we merely investigate the consequences for a design's unconditional properties when applying such a rationale in a mandatory way. 
The described approach should thus be seen as a hypothetical example highlighting the ineffectiveness that can arise from a na\"{\i}ve implementation of a recalculation based on a fixed threshold for 
conditional power.
The corresponding sample size and critical value functions for $m=26$ are depicted in Figure~\ref{fig:binding-recalculation}.
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{latex/figures/naive_adaptive_and_optimal_designs}
    \caption{%
        Original single-stage design; na\"{\i}vely adapted design 
        with sample size $n'(z_m)$ and critical value $c'(z_m)$ functions defined by the conditional optimization problem \eqref{eq:naive-recalc}--\eqref{eq:nmax-naive-recalc} and 
        ${n=79}, {m=26}, {n_{max}=160}, {n_{min}=30}$ and $1-\beta_{cond.}=0.8$;
        the optimal design is the solution of \eqref{eq:optimal-design-start}--\eqref{eq:optimal-design-end}.
    }
    \label{fig:binding-recalculation}
\end{figure}

The mandatory application of the adaptation rule
results in a two-stage design.
The final sample size is thus a random variable $n(Z_m)$.
This implies that the objective criterion of `minimal sample size' is no longer meaningful.
Instead, one could use a weighted sum, $\E[\,n(Z_m)\,] + \eta \sqrt{\V[\,n(Z_m)\,]}$, of the expected value and standard deviation of the sample size as the objective criterion.
The rationale for adding a penalty depending on the standard deviation of the sample size is that it might incur additional costs due to more
complicated logistics (e.g., on demand production of additional drug doses).
For the single-stage design this proposal always reduces to the fixed sample size $n$ since the standard deviation is 0. 
Whether or not the two-stage design is considered better then depends on the choice of $\eta$, 
i.e, the relative weight of the the standard deviation in the objective.
In the particular situation considered here, the expected sample size is $49.5$ and its standard deviation is $29.7$.
Since the original $n$ was $79$, the two-stage
design would be considered `better'
than the single-stage design for $\eta>0.99$.

This comparison, however, ignores the fact that
the original unconditional constraint of and expected power of more than $80\%$
is no longer fulfilled by the design with mandatory
sample size adaptation.
Both expected power ($72.1\%$ \emph{versus} $80.0\%$) and maximal type-I error rate ($2.1\%$ \emph{versus} $2.5\%$) are lower than for the original design.
The difference in operating characteristics is mainly due to the fact
that the new design implicitly introduces a binding early-stopping boundary for futility
when the recalculation problem does not yield a $n'\leq n_{max}$.
This effect is less pronounced when the same recalculation rule is applied 
to an initial two-stage group-sequential design 
which already defines binding early stopping boundaries 
for futility and efficacy.

Similar to suggestions in the literature on binding futility stopping, 
one could modify the initial nominal $\alpha$ and the conditional power threshold $\beta_{cond.}$ until the recalculated design again fulfills the required 
operating characteristics \citep{brannath2004}. 
This approach to define a fully pre-specified two-stage design is, however, unnecessarily complicated in that it 
uses methods like the conditional error principle or a combination function, which are originally intended for \emph{unplanned}
design adaptations.

\section{Optimal pre-planned sample size adaptations}
\label{sec:optimal-adaptation}

Evidently, the na\"{\i}ve application of the predictive power
adaptation rule outlined earlier is needlessly conservative.
Fixing the operating characteristics by tuning the nominal $\alpha$ and
$\beta_{cond.}$ does not address the deeper issue that the sample size and 
critical value functions are still derived based on a merely heuristic recalculation criterion and that 
they still implicitly depend on an original design that is never actually realized via the 
conditional error constraint~\eqref{eq:ce-principle-naive-recalc}.

Instead, the interim time-point $m$, 
the sample size function $n(z_m)$, 
the critical value function $c(z_m)$,
and the early futility- and efficacy boundaries 
($n(z_m)=m$ and $ c(z_m)=\pm\infty$)
can be optimized directly.
For the sample size function $n(z_m)$ \citet{brannath2004} discussed this approach by fitting a polynomial of degree 4 but they did not optimize
any of the other parameters.
In a more general form, \citet{pilz2019} approached the problem using
variational methods.
An \texttt{R} implementation that uses cubic splines for both the $n$ and
$c$ functions is available via the package~\textit{adoptr}~\citep{adoptrjss2020}.

In the following we restrict the considerations to the case
where only expected sample size is of interest and the variation is
ignored, i.e.~$\eta=0$.
The corresponding optimal two-stage design is the solution of
\begin{align}
    \argmin{m, n(\cdot), c(\cdot)} 
        &&                                   \E_\varphi[\,n(Z_m)\,] & \label{eq:optimal-design-start}\\
    \st &&                 \Pr_{0}[\,Z_{n(Z_m)} > c(Z_m)\,] & \leq \alpha \\
        &&     \EP_\varphi\big(n(\cdot), c(\cdot)\big) & \geq 1 - \beta \label{eq:optimal-design-end}
\end{align}
and can be derived numerically using the \texttt{R} package \texttt{adoptr}~\citep{adoptrjss2020}.
Here, $\E_\varphi[\,n(Z_m)\,]$ is the expected sample size under the prior $\varphi$ and expected power needs to be redefined to account for the
fact that $n$ and $c$ are now functions of the interim results
\begin{align}
    \EP_\varphi\big(n(\cdot), c(\cdot)\big) :=
    \Pr_\varphi\big[\,Z_{n(Z_m)}>c(Z_m)\,|\,\Theta\geq0\,\big] \ .
\end{align}
Figure~\ref{fig:binding-recalculation} shows the sample size-, critical value-, and the predictive power function in the example situation discussed earlier together with the single-stage design
and the na\"{\i}ve adaptation based on predictive power discussed in
Section~\ref{sec:naive-adaptations}.
The optimal two-stage design complies with both the predictive power and the maximal type-I error rate constraints and is thus comparable with the original single-stage design in terms of sample size.
The expected sample size is much lower ($56.4$ \emph{versus} $79$) at the cost of a non-zero standard deviation of the expected sample size ($28.5$ \emph{versus} $0$).
Since it is entirely pre-specified, 
the optimal two-stage design does not need to fall back to the conditional error principle to control the maximal type-I error rate.
Instead, it achieves the desired design characteristics in an optimal way since they are incorporated as constraints to problem \eqref{eq:optimal-design-start}--\eqref{eq:optimal-design-end}.
The optimal design fully exhausts the allowable maximal type-I error rate
and complies with the expected power constraint. 
This comes at the cost of a predictive power that can drop as low as 40\% close to the futility boundary
as shown in the right panel of Figure~\ref{fig:binding-recalculation}.
The sample size function of the optimal design is also 
characteristically different from the na\"{\i}ve design.
The latter is convex on the continuation region whereas the optimal
shape has a mode close to the early-futility boundary.
A recalculation based on exceeding a fixed lower threshold for 
some estimator of conditional power always results in a convex
sample size function and can thus never be fully optimal irrespective of how the nominal values of $\alpha$ and $\beta_{cond.}$
are chosen.

The fact that the optimal two-stage design exhibits a monotonously increasing
predictive power rather than a constant predictive power of, say, 80\%
indicates that any recalculation rule keeping predictive power
close to a fixed target value must be inefficient in terms of minimizing expected sample size. 
The mandatory application of the heuristic recalculation rule introduced in Section~\ref{sec:naive-adaptations} would change the optimal design for almost every value of $z_m$ and thus increase expected sample size
of the resulting design.
It may thus appear as if any `recalculation' of an optimal design's sample size using methods 
for unplanned interim analyses was not only 
unnecessary but even counterproductive.
This is, however, only the case if the initial design was optimal and
the planning assumptions still hold.

\section{Consistent unplanned recalculation}
\label{sec:unplanned-adaptation}

The preceding section made the case that the direct optimization of all design parameters is superior to the mandatory application of methods for unplanned design adaptations during the planning phase of a trial.
By definition, 
no optimal design ever needs to be changed as a reaction to trial internal data during the course of the study.
However, there are still situations in which an unplanned sample size recalculation is warranted.
The optimality of a design typically depends on the chosen prior density $\varphi$ and the objective function.
At any point in time, 
the prior density encodes all trial-external evidence 
about the effect size.
This might change over the course of a trial.
For instance, another study might publish new results that trigger a reassessment of the considerations leading to the choice of $\varphi$.

In the following, we discuss two approaches of reacting to such trial-external events in a consistent way 
if the original design minimizes expected sample size. 
Here, `consistent' means that the original optimal design is invariant under
the recalculation rule unless the planning assumptions (prior) are changed or
the time-point of the interim analysis is different from the optimal one.
The considerations are restricted to cases where the result of an adaptation is a final sample size and no further interim analyses are planned for the remainder of the trial.
This means that the unplanned adaptation replaces the pre-planned interim analysis although it might occur at a different point in time. 
All methods can be generalized to an unplanned interim analysis in the
first stage with a subsequent second interim analysis.
This might be necessary when the unplanned interim analysis has to
be conducted shortly after the start of a trial.

To obtain a consistent sample size recalculation rule, 
the conditional error principle can be applied not only to the
maximal type-I error rate but also to the (average) type-II error rate.
Let $m'$ be the sample size of the unplanned interim analysis and
$\psi$ be the probability density function of the revised prior.
We propose to recalculate the new sample size, $n'$ and 
the new critical value $c'$ as the solution of the point-wise optimization problem 
\begin{align}
    \argmin{n',c'} && n' &\\
     \st && \CP(m', n', z_{m'}, c', 0) &\leq \Pr_0[\,Z_{n(Z_m)} > c(Z_m) \cond Z_{m'} = z_{m'}\,] \label{eq:conditional-error-recalc}\\
     && \PP_\psi(m', n', z_{m'}, c') &\geq \nonumber \\ 
     && & \hspace*{-1cm}\Pr_\varphi[\,Z_{n(Z_m)} > c(Z_m) \cond \Theta >0, Z_{m'} = z_{m'}\,] \label{eq:pp-recalc} \ .
\end{align}
Here, constraint~\eqref{eq:pp-recalc} means that the modified design under the new prior has at least as much predictive power as the original design under the original prior. 
Constraint~\eqref{eq:conditional-error-recalc} implements the conditional error principle and thus ensures strict type-I error rate control of the 
procedure. 
The crucial difference to the procedure discussed in Section~\ref{sec:unplanned-adaptation} is the fact that the lower boundary on predictive power varies with $z_m$.

In cases where the interim analysis is conducted at the original $m$, 
the problem simplifies to
\begin{align}
    \argmin{n',c'} && n' & \label{eq:objective-original-interim}\\
    \st 
    && \CP(m, n', z_{m}, c', 0) &\leq \CP(m, n(z_m), z_{m}, c(z_m), 0) \\
    && \PP_\psi(m, n', z_{m}, c') &\geq \PP_\varphi(m, , n(z_m), z_{m}, c(z_m))  \label{eq:pp-original-interim} \ .
\end{align}
Since $\CP$ is monotone in $n'$, the solution is already uniquely 
defined by the constraints, if it exists.
It follows directly that, for $m'=m, \psi=\varphi$, 
the original design ${n'=n(z_{m})}$, ${c'=c(z_m)}$ fulfills both constraints with equality.
The original optimal design is thus indeed invariant under the 
proposed recalculation rule if neither the planning prior not the interim time-point are changed.

Problem \eqref{eq:conditional-error-recalc}--\eqref{eq:pp-recalc} is in no way the only conceivable
means of deriving a consistently recalculated sample size.
For the sake of simplicity, we consider the case of $m'=m$ (i.e.~problem \eqref{eq:objective-original-interim}--\eqref{eq:pp-original-interim}).
Depending on how much the prior changes, 
the required sample size adjustment to meet the predictive power constraint can be quite drastic.
In practice, it might thus be preferable to relax the predictive power constraint in a principled way.
This can be done via the Lagrangian mechanism. 
Let
\begin{align}
    f(n', c') :&= n' \\
    g(n', c') :&= \CP(m, n', z_{m}, c', 0) - \CP(m, n(z_m), z_{m}, c(z_m), 0) \\
    h_\psi(n', c') :&= \PP_\varphi(m, , n(z_m), z_{m}, c(z_m)) - \PP_\psi(m, n', z_{m}, c')
\end{align}
As discussed earlier, the original designs $n$ and $c$ trivially are the unique solution of
\begin{align}
    \argmin{n',c'} && f(n',c') & \\
    \st 
    && g(n',c') &\leq 0 \\
    && h_\varphi(n',c') &\leq 0 \ .
\end{align}
This problem is equivalent to solving the unconstrained problem
\begin{align}
    \argmin{n',c'} f(n',c') - \lambda_g g(n',c') - \lambda_h h_\varphi(n',c') 
\end{align}
for suitable choice of the Lagrange multipliers $\lambda_g$ and $\lambda_h$.
Since the original $n, c$ are optimal, 
$\lambda_g$ and $\lambda_h$ can be obtained directly as solution of
\begin{align}
    0 &= \nabla f(n, c) - \lambda_g \nabla g(n, c) - \lambda_h \nabla h_\varphi(n, c) \\
   \Leftrightarrow (1, 0)^\top &= \big(\nabla g(n, c), \nabla h_\varphi(n, c)\big) \cdot (\lambda_g, \lambda_h)^\top \\
   \Leftrightarrow (\lambda_g, \lambda_h)^\top &= \big(\nabla g(n, c), \nabla h_\varphi(n, c)\big)^{-1} \cdot (1, 0)^\top\ . 
\end{align}
Here, $\nabla$ denotes the gradient as column vector.
The Lagrange multiplier corresponding to the predictive power constraint, $\lambda_h$, encodes the trade-off between predictive power and sample size of the original, 
optimal design at $z_m$.

A recalculated sample size under a different prior $\psi$ can then be obtained by 
keeping $\lambda_h$ fixed and minimizing 
\begin{align}
    \argmin{n',c'} && n'&-\lambda_h\, h_\psi(n',c') \\
     \st && g(n', c') &\leq 0 \ . 
\end{align}
This effectively relaxes the predictive power constraint by allowing the same trade-off 
between sample size and predictive power that was implied by the original optimal design.
The conditional type-I error rate constraint is not relaxed to comply with the 
conditional error principle.
This approach is closely related to the formulation of the optimal-design problem given in~\citet{jennison2015} and the utility perspective on sample 
size derivation presented in~\citep{kunzmann2020}.
In contrast to the situation analyzed by \citet{jennison2015}, however,
$\lambda_h$ does depend on $z_m$.
This is due to the fact that expected sample size is formed with respect to the unconditional
prior while expected power uses the prior conditional on $\Theta>0$.

We illustrate both approaches by example of the optimal two-stage design that minimizes the objective 
$\E_\varphi[\,n(Z_m)\,]$ subject to a maximal type-I error rate of $0.025$ and a minimal expected power of $0.8$, see~Figure~\ref{fig:binding-recalculation}.  
This design has a pre-planned interim analysis after $m=35$ individuals.
Now assume that
the prior $\varphi$ needs to be revised due to new trial-external 
information at the pre-planned interim analysis.
We consider members of the class of shifted truncated Normal distributions 
\begin{align}
    \psi_{\mu}(\theta) := \frac{\displaystyle \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\phi\left(\frac{\theta - \mu}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - \mu}{0.2}\right) - \Phi\left(\frac{-0.5 - \mu}{0.2}\right)}
\end{align}
as revised priors, i.e. the original prior~$\varphi=\psi_{0.4}$.
Figure~\ref{fig:recalculation-old-m} shows the behavior of the recalculated sample size and critical value in this scenario for observed effects of $z_m/\sqrt{m}=0.25, 0.3, 0.35$ and varying mean of the revised prior if the originally planned interim time-point is maintained.
\begin{figure}
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-type-II-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect.
        }
        \label{fig:recalculation-old-m}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-type-II-earlier}
        \caption{%
            Earlier-than-planned interim analysis at an observed effect of $0.3$ and varying interim time-point.
        }
        \label{fig:recalculation-earlier}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-lambda-approach-original-m}
        \caption{%
            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect and the $\lambda$ approach.
        }
        \label{fig:recalculation-lambda}
    \end{subfigure}
    \caption{%
        Comparison of consistent unplanned recalculation schemes.
    }
\end{figure}
Clearly, the original design is indeed invariant if the prior remains unchanged.
The sample size is increased if the new prior is more conservative
than $\varphi$ and \textit{vice~versa}.

Of course, it could also be necessary to conduct the interim analysis earlier or later than originally anticipated, at $m'\neq m$. 
In this case, the design is no longer invariant since the earlier 
or later time point already implies a deviation from the 
original optimal design even if the prior remains the same.
Figure~\ref{fig:recalculation-earlier} shows the resulting adaptations
for earlier-than-planned time points of the interim analysis 
while the observed effect is kept fixed at $0.3$.
The formula for the conditional error in this situation is given in the Appendix.

Figure~\ref{fig:recalculation-lambda} shows the recalculated sample size
for $m'=m=35$ and the $\lambda$-approach. 
The original design is again (approximately) invariant under recalculation.
Compared to the fixed conditional type-II error approach, predictive power 
under the $\lambda$-approach is more flexible and might drop below the predictive power under the original design, see Figure~\ref{fig:recalculation-lambda}, right panel.
This flexibility in terms of predictive power leads to a much smaller deviation from the original design's sample size.
The $\lambda$-approach transitions smoothly from 
an increase of the original sample size due to a slightly more conservative prior to a decrease under very conservative priors.
This behavior leads to less extreme adjustments and is thus 
more realistic in practice.
After all, a slight loss in (predictive) power as compared to the
original design might be tolerable if the sample size can be kept at
a moderate level in exchange. 
Of course, it is still possible to stop the trial early for futility
at any point without compromising type one error rate control 
if the predictive power is considered too low to warrant a continuation.

\section{Discussion} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Monitoring the power of the remainder of an ongoing trial constitutes an estimation problem since the conditional power
is a function of the unknown treatment effect. 
An often overlooked aspect of this estimation problem is that it is 
conditional on a positive treatment effect since the notion of power is conditional on there being an non-null effect~\citep{kunzmann2020}.
In many practical situations where the prior mass is concentrated on 
non-null effects the effect of conditioning is negligible but its importance increases with the vagueness of a the prior. 
This lack of conditioning on a positive treatment effect is one of the reasons observed conditional power performs rather poorly as an estimator of the unknown conditional power.
Assumed conditional power for $\theta_1>0$ does reflect this conditionality.
It is, however,
merely a special case of predictive power with a 
point prior on the effect size.
In situations where the effect size is fairy well known, a point prior might constitute an acceptable, simpler approximation.
In any other case, the Bayesian predictive power allows \textit{a~priori} information to be incorporated in a more fine-grained way and
guarantees optimal mean-squared-error performance.
If an unconditional measure for the probability to reject the null hypothesis is sought, 
a conditional version of the probability of success (or assurance) can be derived in an analogue way~\citep{kunzmann2020}.

Great care should be taken when predictive power 
or another estimator of conditional power
is used to modify an ongoing trial's sample size.
Altering a simple starting design with a seemingly intuitive recalculation rule 
can actually lead to overall \emph{less effective} trials.
Techniques originally intended for \emph{unplanned} adaptations of trials 
should not be used for a \emph{pre-planned} sample size recalculation.
Instead, an optimal two-stage design should directly be derived 
for the objective criterion of interest.
By definition, no trial-internal event can then justify a sample size recalculation.
Only trial external events, such as a change in the objective criterion,
the emergence of new trial-external evidence,
or unforeseen deviations from the planned interim analysis time-point
may require a reassessment of the sample size of an optimal design.
A generic recalculation rule might lead to the paradoxical situation that even an optimal
starting design is always modified during the interim analysis 
- even if the planning assumptions remain unaltered throughout the trial.
This is clearly ineffective and, consequently, 
an optimal starting design should be invariant under a sensible adaptation rule 
if the planning assumptions remain unchanged.
For instance, this minimal consistency property is not fulfilled when recalculating a design's
sample size based on a fixed threshold for its minimal conditional power 
since designs optimizing expected sample size tend to have variable conditional power.

We propose two consistent approaches to adjusting an ongoing optimal design to 
newly emerging trial-external data.
The first method applies the conditional error principle to the (average) type-II error rate in a 
similar way to its use in controlling the maximal type-I error rate.
The method is easy to implement and consistent in the above defined way.
We would like to stress that this is by no means the only way of 
conducting a consistent sample size recalculation.
The alternative $\lambda$-approach allows a trade-off between predictive power 
and sample size leading to smaller sample size deviations from the original design.
This comes at the risk of potentially loosing predictive power.
Both methods differ by the quantity that is held fixed during the recalculation.
For the first method, the predictive power under the old design and the old prior
is maintained for the new design under the new prior whereas the $\lambda$-approach
only keeps the trade-off between predictive power and sample size fixed but allows the
actual predictive power to deviate form the original design.

These considerations show how absolutely crucial the initial planning stage of a trial is. 
Adaptive methods should not be taken as an excuse to start with a
sub-optimal design and rely on a later sample size recalculation.
Ideally, all uncertainty is quantified during the planning phase to 
the best possible extend and integrated in the initial design via a 
planning prior.
This does not mean that a complex design is always the best choice.
If the variability of the final sample size and the operational 
burden of conducting interim analyses is penalized strong enough in
the objective criterion, a simple one-stage design might very well
perform better than more complex alternatives.
Also, 
optimal-two stage designs require optimizing over function-spaces to find the optimal sample size and critical value functions.
Obtaining a stable solution might thus be hard in practice and
group-sequential designs are a viable approximation since optimizing the stage-wise sample sizes and stopping boundaries only requires optimizing over a small set of real parameters.
It is well-known, that optimal group-sequential designs approximate the performance of optimal two-stage designs with variable sample sizes sufficiently well~\citep{wassmer2016}.
Still, the same principle considerations apply: 
An optimal (group-sequential) design only needs to be revised if either the objective function or the underlying planning assumptions, i.e., the prior on the effect size, changes.

Adaptations of initially sub-optimal designs are another potential 
application of the new recalculation scheme. 
Clearly, the single-stage design with $n=79$ is not minimizing 
\emph{expected} sample size. 
To switch to a near-optimal adaptive design mid-trial, 
one could derive the optimal two-stage design given the current best prior \textit{post-hoc} and 
then use the $\lambda$-approach to recalculate the initial design while controlling the maximal 
conditional type-I error rate in the usual way.
The resulting design will not be optimal from an unconditional perspective since the conditional error function of the initial, sub-optimal design has to be respected but the power-sample size trade-off during the interim analysis 
will be similar to the optimal design due to the choice of $\lambda_h$.

A limitation of the proposed recalculation methods is the fact that they 
rely on the conditional error principle for strict type-I-error rate control.
The conditional error principle can be difficult to extend to cases with nuisance
parameters \citep{gutjahr2011}.
In practice, a simple plug-in approach similar to the approach in blinded sample size reassessment might be viable but have not yet been investigated more thoroughly.
Alternatively, any other form of pre-specifying a combination function for the stage-wise p values might be used to control the type-I-error rate.
This combination function should then also be optimized over during the planning stage 
to avoid inconsistencies. 
In particular, the optimal combination function of a two stage-design minimizing expected sample size can be approximated with an inverse normal combination test~\citep{pilz2019}.

\bibliographystyle{plainnat}
\bibliography{references.bib}

\section*{Appendix}

The conditional error for an earlier-than-planned interim analysis ($m'<m$)
is 
\begin{align}
    & \Pr_\theta\big[\,Z_{n(Z_m)} > c(Z_m) \cond Z_{m'} = z_{m'}\,\big] \\
    =& \int \CP_\theta\big(m,n(z_m),z_m, c(z_m), \theta\big)\  p_\theta(z_m \cond z_{m'}) \operatorname{d} z_m
\end{align}
where $p_\theta(z_m \cond z_{m'})$ denotes the probability density function of the outcome at $m$ given the partial observed outcome up to $m'$ (see equation~\eqref{eq:joint-distribution}).

\end{document}

### 2020-10-13 latex/manuscript.tex
Updates from Overleaf

@@
     \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk} \\
-
     Michael J. Grayling \\
@@
     \texttt{michael.grayling@newcastle.ac.uk} \\
-
     Kim May Lee \\
@@
     \texttt{k.m.lee@qmul.ac.uk} \\
-
     David S.\ Robertson \\
@@
     \texttt{david.robertson@mrc-bsu.cam.ac.uk} \\
-
     Kaspar Rufibach \\
@@
     \texttt{kaspar.rufibach@roche.com} \\
-
     James M. S. Wason \\
@@
         \centering
-        \includegraphics[width=.9\textwidth]{latex/figures/acp_ocp_pp_sampling_distributions}
+        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_sampling_distributions}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{latex/figures/acp_ocp_pp_bias_mae_mse}
+        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_bias_mae_mse}
         \caption{%
@@
     \centering
-    \includegraphics[width=\textwidth]{latex/figures/naive_adaptive_and_optimal_designs}
+    \includegraphics[width=\textwidth]{figures/naive_adaptive_and_optimal_designs}
     \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-type-II-original-m}
+        \includegraphics[width=.9\textwidth]{figures/recalculation-type-II-original-m}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-type-II-earlier}
+        \includegraphics[width=.9\textwidth]{figures/recalculation-type-II-earlier}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{latex/figures/recalculation-lambda-approach-original-m}
+        \includegraphics[width=.9\textwidth]{figures/recalculation-lambda-approach-original-m}
         \caption{%

### 2020-10-13 .binder/install.R
lineendings

@@ -1,5 +1,5 @@
-install.packages("devtools")
 install.packages("tidyverse")
+install.packages("tinytex")
 install.packages("patchwork")
 install.packages("truncnorm")
 install.packages("adoptr")

### 2020-10-13 latex/manuscript.tex
lineendings
...
Updates from Overleaf

@@
-\documentclass{article}
+\documentclass[12pt]{article}
 
@@
 \usepackage{color}
-\usepackage{natbib}
+\usepackage[numbers,super,sort&compress]{NJDnatbib}
+\bibliographystyle{WileyNJD-AMA}
 \usepackage{scalerel}
@@
 derivation takes a Bayesian
-view on determining the initial sample size of a clinical trial~\citep{spiegelhalter1994}.
+view on determining the initial sample size of a clinical trial~\cite{spiegelhalter1994}.
 The Bayesian component requires the specification of an informative
@@
 to recalculate the initial sample size of a design based on data observed
-within the trial itself, see e.g.~\citet{bauer2016}.
+within the trial itself~\cite{bauer2016}.
 The rationale behind this approach is to use the accruing evidence about the unknown parameters driving the sample size derivation to
@@
 or not.
-Methods for the latter `blinded interim analysis' are particularly useful if relevant nuisance parameters such as the variance are also unknown \citep{birkett1994internal,bauer2016}.
+Methods for the latter `blinded interim analysis' are particularly useful if relevant nuisance parameters such as the variance are also unknown \cite{birkett1994internal,bauer2016}.
 We focus on the unblinded case which allows for a more precise
@@
 The maximal type-I error rate constraint is then usually protected by
-applying the conditional error principle~\citep{muller2004,brannath2012}
+applying the conditional error principle~\cite{muller2004,brannath2012}
 or an equivalent formulation via $\alpha$-spending or
-combination functions~\citep{bauer2016}.
+combination functions~\cite{bauer2016}.
 Often the sample size of the current
@@
 interim analysis again exceeds the initial threshold for unconditional
-power~\citep{proschan1995}.
+power~\cite{proschan1995}.
 Any recalculation based on conditional power arguments must
@@
 The latter `plug-in' approach is often criticized for
-`... [using] the interim estimate of the effect [...] twice ...' \citep[p.~330]{bauer2016} (see also~\cite{bauer2006}).
+`... [using] the interim estimate of the effect [...] twice ...' Bauer~\textit{et~al.},~p.~330~\cite{bauer2016} and \cite{bauer2006}.
 Thirdly,
@@
 Within the hybrid Bayesian framework this is usually referred to as
-`predictive power'~\citep{spiegelhalter1994,bauer2016}.
+`predictive power'~\cite{spiegelhalter1994,bauer2016}.
 
@@
 Approaches to defining such a power constraint under
-different assumptions were reviewed in~\citet{kunzmann2020}.
+different assumptions were reviewed in~\cite{kunzmann2020}.
 
@@
 Here the conditioning on a positive effect is crucial to justify the notion
-of `power' in contrast to the unconditional `probability of success'~\citep{kunzmann2020}.
+of `power' in contrast to the unconditional `probability of success'~\cite{kunzmann2020}.
 
@@
 less commonly taken in the literature but provides a consistent framework to
-compare the characteristics of different methods 
-(see also~\citep[Discussion]{bauer2006}).
+compare the characteristics of different methods~\cite{bauer2006}.
 
@@
 Secondly, the observed effect $\widehat{\theta}_m$ can be used
-as a plug-in estimator for the unknown effect size
-(see, e.g.~\citep{proschan1995}).
+as a plug-in estimator for the unknown effect size~\cite{proschan1995}.
 This quantity is sometimes referred to as `observed conditional power'
@@
 \end{align}
-differs slightly from the one proposed by~\citet{spiegelhalter1994}
+differs slightly from the one proposed by~\cite{spiegelhalter1994}
 in that we condition on a positive effect size $\Theta>0$.
@@
 the location and vagueness (variance) of the prior.
-Following~\citep{kunzmann2020} the required 
+Following~\cite{kunzmann2020} the required
 sample size $n$ is then determined by requiring
@@
 Also, the negligence of the sampling variation of $\widehat{\theta}_m$, which $\OCP$ plugs in the expression for conditional power, leads to a larger variance for intermediate values of $\theta$ and the characteristic
-U-shape \citep{bauer2006}.
+U-shape \cite{bauer2006}.
 This high variance of $\OCP$ directly translates to a relatively large mean absolute error (MAE) and mean squared error (MSE) when estimating conditional power,
@@
 Monitoring the predictive power of an ongoing trial naturally raises the question of whether this information can be leveraged to improve the operating characteristics of a trial.
-Typically, sample size recalculation is considered in the context of group-sequential designs at the pre-specified interim analysis \citep{proschan1995,brannath2004,bauer2016}.
+Typically, sample size recalculation is considered in the context of group-sequential designs at the pre-specified interim analysis \cite{proschan1995,brannath2004,bauer2016}.
 From a statistical perspective,
@@
 Different choices for the estimator of conditional power
-and the choice of $\beta_{cond.}$ have been discussed in the literature \citep{proschan1995,brannath2004,bauer2016}.
-Strict overall type-I error rate control can be maintained by invoking the conditional error principle \citep{muller2001,muller2004,brannath2012}, i.e., by limiting the maximal type-I error rate of the new design to the maximal conditional type-I error rate of the original design given the observed $z_m$.
+and the choice of $\beta_{cond.}$ have been discussed in the literature \cite{proschan1995,brannath2004,bauer2016}.
+Strict overall type-I error rate control can be maintained by invoking the conditional error principle \cite{muller2001,muller2004,brannath2012}, i.e., by limiting the maximal type-I error rate of the new design to the maximal conditional type-I error rate of the original design given the observed $z_m$.
 Often, trial protocols leave the exact choice of $1-\beta_{cond.}$ open since it can be chosen \textit{ad~hoc}
 without compromising strict type-I-error rate control.
-The basic concept of readjusting the sample size to achieve the desired conditional power is, however, a common approach, see for instance,~\citet{bhatt2013} and~\citet{mehta2009}.
+The basic concept of readjusting the sample size to achieve the desired conditional power is, however, a common approach, see for instance,~\cite{bhatt2013} and~\cite{mehta2009}.
 
@@
 one could modify the initial nominal $\alpha$ and the conditional power threshold $\beta_{cond.}$ until the recalculated design again fulfills the required
-operating characteristics \citep{brannath2004}. 
+operating characteristics \cite{brannath2004}.
 This approach to define a fully pre-specified two-stage design is, however, unnecessarily complicated in that it
@@
 can be optimized directly.
-For the sample size function $n(z_m)$ \citet{brannath2004} discussed this approach by fitting a polynomial of degree 4 but they did not optimize
+For the sample size function $n(z_m)$ \cite{brannath2004} discussed this approach by fitting a polynomial of degree 4 but they did not optimize
 any of the other parameters.
-In a more general form, \citet{pilz2019} approached the problem using
+In a more general form, \cite{pilz2019} approached the problem using
 variational methods.
 An \texttt{R} implementation that uses cubic splines for both the $n$ and
-$c$ functions is available via the package~\textit{adoptr}~\citep{adoptrjss2020}.
+$c$ functions is available via the package~\textit{adoptr}~\cite{adoptrjss2020}.
 
@@
 \end{align}
-and can be derived numerically using the \texttt{R} package \texttt{adoptr}~\citep{adoptrjss2020}.
+and can be derived numerically using the \texttt{R} package \texttt{adoptr}~\cite{adoptrjss2020}.
 Here, $\E_\varphi[\,n(Z_m)\,]$ is the expected sample size under the prior $\varphi$ and expected power needs to be redefined to account for the
@@
 conditional error principle.
-This approach is closely related to the formulation of the optimal-design problem given in~\citet{jennison2015} and the utility perspective on sample 
-size derivation presented in~\citep{kunzmann2020}.
-In contrast to the situation analyzed by \citet{jennison2015}, however,
+This approach is closely related to the formulation of the optimal-design problem given in~\cite{jennison2015} and the utility perspective on sample
+size derivation presented in~\cite{kunzmann2020}.
+In contrast to the situation analyzed by \cite{jennison2015}, however,
 $\lambda_h$ does depend on $z_m$.
@@
 An often overlooked aspect of this estimation problem is that it is
-conditional on a positive treatment effect since the notion of power is conditional on there being an non-null effect~\citep{kunzmann2020}.
+conditional on a positive treatment effect since the notion of power is conditional on there being an non-null effect~\cite{kunzmann2020}.
 In many practical situations where the prior mass is concentrated on
@@
 If an unconditional measure for the probability to reject the null hypothesis is sought,
-a conditional version of the probability of success (or assurance) can be derived in an analogue way~\citep{kunzmann2020}.
+a conditional version of the probability of success (or assurance) can be derived in an analogue way~\cite{kunzmann2020}.
 
@@
 group-sequential designs are a viable approximation since optimizing the stage-wise sample sizes and stopping boundaries only requires optimizing over a small set of real parameters.
-It is well-known, that optimal group-sequential designs approximate the performance of optimal two-stage designs with variable sample sizes sufficiently well~\citep{wassmer2016}.
+It is well-known, that optimal group-sequential designs approximate the performance of optimal two-stage designs with variable sample sizes sufficiently well~\cite{wassmer2016}.
 Still, the same principle considerations apply:
@@
 The conditional error principle can be difficult to extend to cases with nuisance
-parameters \citep{gutjahr2011}.
+parameters \cite{gutjahr2011}.
 In practice, a simple plug-in approach similar to the approach in blinded sample size reassessment might be viable but have not yet been investigated more thoroughly.
@@
 to avoid inconsistencies.
-In particular, the optimal combination function of a two stage-design minimizing expected sample size can be approximated with an inverse normal combination test~\citep{pilz2019}.
+In particular, the optimal combination function of a two stage-design minimizing expected sample size can be approximated with an inverse normal combination test~\cite{pilz2019}.
 
-\bibliographystyle{plainnat}
-\bibliography{references.bib}
+\bibliography{manuscript.bib}
 

### 2020-10-13 latex/manuscript.tex
Updates from Overleaf

@@
 \newcommand{\kr}[1]{{\color{red}#1}}
+\usepackage{comment}
 
@@
 
+%\begin{comment}
 \author{
@@
         Cambridge CB2 0SR,
-        United Kingdom
+        United Kingdom,
+        \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk}
     }\\
@@
     University of Cambridge \\
-    \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk} \\
     Michael J. Grayling \\
@@
     Newcastle University \\
-    \texttt{michael.grayling@newcastle.ac.uk} \\
     Kim May Lee \\
@@
     Queen Mary University of London\\
-    \texttt{k.m.lee@qmul.ac.uk} \\
     David S.\ Robertson \\
@@
     University of Cambridge\\
-    \texttt{david.robertson@mrc-bsu.cam.ac.uk} \\
     Kaspar Rufibach \\
@@
     F. Hoffmann-La Roche, Basel\\
-    \texttt{kaspar.rufibach@roche.com} \\
     James M. S. Wason \\
@@
      MRC Biostatistics Unit \\
-    University of Cambridge \\
-    \texttt{james.wason@newcastle.ac.uk}
+    University of Cambridge
 }
+%\end{comment}
+%\author{}
 
@@
 \begin{abstract}
-Determining the sample size of a confirmatory trial is challenging.
-If the sample size is too small, the chance of detecting an effect is small.
-If it is too big, trial duration and costs are needlessly inflated.
 Adapting the final sample size of a trial to the evidence accruing
-during the trial is thus a natural way to address
-this planning uncertainty.
-In a regulatory environment,
-designs with adaptive sample size need to
+during the trial is a natural way to address planning uncertainty.
+Designs with adaptive sample size need to
 account for their optional stopping to guarantee
@@
 after unplanned changes of the initial sample size have been proposed
-in the literature rendering interim analyses for
-the purpose of sample size recalculation feasible in
-practice.
-Any form of interim analysis allows the analysis of
-a trial's prospects to finally reject the null hypothesis
-given the data observed up to the interim analysis.
+in the literature.
+This makes interim analyses for the purpose of sample size 
+recalculation feasible in a regulatory context.
 Since the sample size is usually determined via an
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_example}
+        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_example}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_sampling_distributions}
+        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_sampling_distributions}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figures/acp_ocp_pp_bias_mae_mse}
+        \includegraphics[width=.85\textwidth]{figures/acp_ocp_pp_bias_mae_mse}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figures/recalculation-type-II-original-m}
+        \includegraphics[width=.8\textwidth]{figures/recalculation-type-II-original-m}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figures/recalculation-type-II-earlier}
+        \includegraphics[width=.8\textwidth]{figures/recalculation-type-II-earlier}
         \caption{%
@@
         \centering
-        \includegraphics[width=.9\textwidth]{figures/recalculation-lambda-approach-original-m}
+        \includegraphics[width=.8\textwidth]{figures/recalculation-lambda-approach-original-m}
         \caption{%

### 2020-10-13 R/04-consistent-recalculation-lambda-approach.R
fixed bug

@@ -67,67 +67,6 @@ optimal_design <- minimize(
 )$design
 
 
-
-# Determine 'Lagrange-multiplier' lambda ---------------------------------
-# function to determine optimal design given lambda
-f <- function(lambda) {
-    minimize(
-        composite({ESS - lambda*EP}),
-        subject_to(
-            TOER <= alpha
-        ),
-        get_initial_design(
-            theta = theta1,
-            alpha = alpha,
-            beta  = beta,
-            type  = 'two-stage',
-            dist  = Normal(two_armed = FALSE),
-            order = 5L # reduced precision, we only neeed to get a good proxy for the power to match, faster!
-        ),
-        opts = list(algorithm = "NLOPT_LN_COBYLA", xtol_rel = 1e-05, maxeval = 5e4)
-    )$design
-}
-# evaluate on grid (smoothness unknown!) to pick lambda that gives
-# target power of 1 - beta
-tbl_lambda <- tibble(
-        lambda = seq(200, 300, by = 5),
-        design = map(lambda, ~future({f(.)})) # spawn futures
-    ) %>%
-    mutate(
-        design = map(design, value), # collect futures
-         delta = map_dbl(design, ~abs(evaluate(EP, .) - (1 - beta)))
-    )
-# diagnostic plot, not saved
-tbl_lambda %>%
-    ggplot() +
-    aes(lambda, delta) +
-    geom_line() +
-    geom_point() +
-    ylab("power difference")
-cat("original optimal design:\n\r")
-summary(
-        optimal_design,
-        `E[N]` = ESS,
-        `SD[N]` = SDSS,
-        `MTOER` = TOER,
-        `Expected Power` = EP
-    ) %>%
-    show()
-# determine optimal lambda
-lambda <- tbl_lambda%>% arrange(delta) %>% pull(lambda) %>% head(1)
-cat(sprintf("matched-lambda optimal design, lambda=%.2f:\n\r", lambda))
-tbl_lambda %>%
-    arrange(delta) %>%
-    pull(design) %>%
-    .[[1]] %>%
-    summary(
-         `E[N]` = ESS,
-        `SD[N]` = SDSS,
-        `MTOER` = TOER,
-        `Expected Power` = EP
-    ) %>%
-    show()
-
 # Plot adaptations under lambda rule -------------------------------------------
     m <- 35
 n_max <- 160
@@ -165,7 +104,7 @@ adapt_lambda <- function(priormean, zm) {
[…]

### 2020-10-13 latex/manuscript.tex
...
added code links

@@
 \usepackage{color}
+\usepackage{hyperref}
 \usepackage[numbers,super,sort&compress]{NJDnatbib}
@@
 
-%\begin{comment}
+\begin{comment}
 \author{
@@
 }
-%\end{comment}
-%\author{}
+\end{comment}
+\author{}
 
@@
 
+\subsection*{Data Availibility Statement}
+Data sharing is not applicable to this article as no datasets were generated and
+the research is entirely theoretical.
+The source code is available at
+\href{https://github.com/kkmann/unblinded-sample-size-adaptation}{github.com}
+and
+\href{https://doi.org/10.5281/zenodo.3925752}{zenodo.org}.
+
 \bibliography{manuscript.bib}

### 2021-08-13 latex/manuscript.tex
...
major refactor

--- 2020-10-13 latex/manuscript.tex
@@
 \newcommand{\V}{\operatorname{V}}
+\newcommand{\solve}[1]{\operatornamewithlimits{solve:\ }_{#1}}
 \newcommand{\st}{\operatorname{subject\ to:\ }}
@@
 \newcommand{\argmax}[1]{\operatornamewithlimits{argmax:\ }_{#1}}
+\newcommand{\etal}{\textit{et~al.}}
 
@@
 
-%\begin{comment}
 \author{
-    Kevin Kunzmann \thanks{
-        MRC Biostatistics Unit, University of Cambridge,
-        Cambridge Institute of Public Health,
-        Forvie Site, Robinson Way,
-        Cambridge Biomedical Campus,
-        Cambridge CB2 0SR,
-        United Kingdom,
-        \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk}
-    }\\
-    MRC Biostatistics Unit \\
-    University of Cambridge \\
-    Michael J. Grayling \\
-    Population Health Sciences Institute \\
-    Newcastle University \\
-    Kim May Lee \\
-    Pragmatic Clinical Trials Unit\\
-    Queen Mary University of London\\
-    David S.\ Robertson \\
-    MRC Biostatisics Unit\\
-    University of Cambridge\\
-    Kaspar Rufibach \\
-    Methods, Collaboration, and Outreach Group (MCO) \\
-    Department of Biostatistics\\
-    F. Hoffmann-La Roche, Basel\\
-    James M. S. Wason \\
-    Population Health Sciences Institute \\
-    Newcastle University\\
-    and \\
-     MRC Biostatistics Unit \\
-    University of Cambridge
+    \footnotesize Kevin Kunzmann\\
+    \footnotesize MRC Biostatistics Unit, University of Cambridge \\
+    \footnotesize
+    East Forvie Building, Robinson Way,
+    Cambridge Biomedical Campus \\
+    \footnotesize Cambridge CB2 0SR,
+    United Kingdom \\
+    \footnotesize \texttt{kevin.kunzmann@mrc-bsu.cam.ac.uk}
+    \and
+    \footnotesize Michael J. Grayling \\
+    \footnotesize Population Health Sciences Institute, Newcastle University
+    \and
+    \footnotesize Kim May Lee \\
+    \footnotesize Institute of Psychiatry, Psychology and Neuroscience, King's College London
+    \and
+    \footnotesize David S.\ Robertson \\
+    \footnotesize MRC Biostatisics Unit, University of Cambridge
+    \and
+    \footnotesize Kaspar Rufibach \\
+    \footnotesize Methods, Collaboration, and Outreach Group (MCO) \\
+    \footnotesize Department of Biostatistics, F. Hoffmann-La Roche, Basel
+    \and
+    \footnotesize James M. S. Wason \\
+    \footnotesize Population Health Sciences Institute, Newcastle University and \\
+    \footnotesize MRC Biostatistics Unit, University of Cambridge
 }
-%\end{comment}
-%\author{}
 
@@
 during the trial is a natural way to address planning uncertainty.
-Designs with adaptive sample size need to
-account for their optional stopping to guarantee
-strict type-I error-rate control.
-A variety of different methods to maintain type-I error-rate control
-after unplanned changes of the initial sample size have been proposed
-in the literature.
-This makes interim analyses for the purpose of sample size 
-recalculation feasible in a regulatory context.
-Since the sample size is usually determined via an
+Since the sample size is usually determined by an
 argument based on the power of the trial,
@@
 how the final sample size should be determined conditional on the accrued information.
-Conditional power is a concept often put forward in this context.
-Since it depends on the unknown effect size,
-we take a strict estimation perspective and compare
-assumed conditional power, observed conditional power,
-and predictive power with respect to their properties as estimators
-of the unknown conditional power.
-We then demonstrate that pre-planning an interim analysis using
-methodology for unplanned interim analyses is ineffective and
-naturally leads to the concept of optimal two-stage designs.
-We conclude that unplanned design adaptations should only be conducted
+To this end, we first review and compare common approaches to estimating conditional power
+which is often used in heuristic sample size recalculation rules.
+We then discuss the connection of heuristic sample size recalculation and
+optimal two-stage designs demonstrating that the latter is the superior approach in a fully pre-planned setting.
+Hence, unplanned design adaptations should only be conducted
 as reaction to trial-external new evidence, operational needs to violate the originally chosen design,
-or \textit{post~hoc} changes in the objective criterion.
-Finally, we show that commonly discussed sample size recalculation rules can lead to paradoxical outcomes and propose two alternative ways of reacting to
-newly emerging trial-external evidence.
+or \textit{post~hoc} changes in the optimality criterion but not as a reaction to trial-internal data.
+We are able to show that commonly discussed sample size recalculation rules lead to paradoxical adaptations
+where an initially planned optimal design is not invariant under the adaptation rule
+even if the planning assumptions do not change.
+Finally, we propose two alternative ways of reacting to
+newly emerging trial-external evidence in ways that are consistent with the originally
+planned design to avoid such inconsistencies.
 \end{abstract}
@@
 parameters underlying the hypothesis of interest.
-Often the alternative hypothesis is the superiority of a new intervention arm
-over a treatment-as-usual control arm and is formalized via
-the average treatment effect, e.g., the difference of means between the
-two randomly allocated treatment arms.
 Clinical data collection is expensive and time consuming leading to a
@@
 The conventional statistical criteria to determine the sample size of a trial
-are a one-sided type-I error rate of
+are a one-sided type~I error rate of
 2.5\% and a power of 80\% or 90\%.
-Since power is a function of the unknown effect size the initial design
-must be specified under substantial uncertainty about the magnitude of the effect which the trial hopes to detect eventually.
+Since power is a function of the unknown effect size, the initial design
+must be specified under substantial uncertainty.
 Mainly, two ways of addressing this challenge have been put forward in the literature.
 
-Firstly, the so-called `hybrid' approach to sample size
+Firstly, the so-called "hybrid" approach to sample size
 derivation takes a Bayesian
-view on determining the initial sample size of a clinical trial~\cite{spiegelhalter1994}.
-The Bayesian component requires the specification of an informative
+view on determining the initial sample size of a clinical trial~\cite{spiegelhalter1994},
+which requires the specification of an informative
 prior on the parameters of interest.
-This then allows reasoning about the `expected power' of a trial as a
+This then allows reasoning about the "expected power" of a trial as a
 function of its sample size and, consequently, determination of the sample size
@@
 This concept is a straight-forward extension of the usual practice of
-computing the power under a fixed point alternative, which can easily be recovered as a special case when considering a point prior.
+computing the power under a fixed point alternative, which can be recovered as a special case when considering a point prior.
 The advantage lies in the fact that the \textit{a~priori} information about the magnitude of the effect size is faithfully
 reflected in the sample size derivation.
-Since the actual analysis is still frequentist, type-I error rate control is not compromised.
+Since the actual analysis is still frequentist, type~I error rate control is not compromised.
 
@@
 The rationale behind this approach is to use the accruing evidence about the unknown parameters driving the sample size derivation to
-`correct' the sample size mid-trial.
+"correct" the sample size mid-trial.
 During the interim analysis, the accrued data can either be unblinded
 or not.
-Methods for the latter `blinded interim analysis' are particularly useful if relevant nuisance parameters such as the variance are also unknown \cite{birkett1994internal,bauer2016}.
+Methods for the latter blinded interim analysis are particularly useful if relevant nuisance parameters such as the variance are also unknown \cite{birkett1994internal,bauer2016}.
 We focus on the unblinded case which allows for a more precise
@@
 the blind.
-The maximal type-I error rate constraint is then usually protected by
+The maximal type~I error rate constraint is then usually protected by
 applying the conditional error principle~\cite{muller2004,brannath2012}
@@
 just as unconditional power,
-depends on the unknown underlying parameters.
-It must thus be
+depends on the unknown underlying parameters
+and must thus be
 estimated to inform a sample size recalculation.
@@
 
-Firstly,
-`conditional power' is the probability to reject the null hypothesis given the interim data as a function of the unknown parameters.
-In a slight abuse of terminology, the same term is often used to refer to
+Conditional power is the probability to reject the null hypothesis given the interim data as a function of the unknown parameters.
+Firstly, in a slight abuse of terminology, the same term is often used to refer to
 the assumed conditional power that is obtained by plugging in the
@@
 Secondly,
-authors have proposed to use the `observed conditional power' instead, which replaces the unknown parameters with their maximum-likelihood estimates.
-The latter `plug-in' approach is often criticized for
-`... [using] the interim estimate of the effect [...] twice ...' Bauer~\textit{et~al.},~p.~330~\cite{bauer2016} and \cite{bauer2006}.
+authors have proposed to use the "observed conditional power" instead, which replaces the unknown parameters with their maximum-likelihood estimates.
+The latter approach is often criticized for
+`... [using] the interim estimate of the effect [...] twice ...' Bauer~\textit{et~al.},~p.~330~\cite{bauer2016} and Bauer and K\"onig\cite{bauer2006}.
 Thirdly,
@@
 Within the hybrid Bayesian framework this is usually referred to as
-`predictive power'~\cite{spiegelhalter1994,bauer2016}.
+"predictive power"~\cite{spiegelhalter1994,bauer2016}.
 
 The purpose of this paper is to argue that pre-planned sample size adaptations
-using methodology intended for unplanned interim analyses to react to within-trial interim data are inefficient and unnecessary if the original design was planned optimally.
-We then discuss cases where an unplanned design adaptation might still be warranted and propose two criteria that allow such an adaptation that is
-consistent with the original design.
-We first review assumed \mbox{conditional-,} observed \mbox{conditional-,} and predictive power
+using methodology intended for unplanned interim analyses to react to within-trial interim data are inefficient and unnecessary if the original design was planned optimally and should thus be avoided.
+We discuss cases where an unplanned design adaptation might still be warranted as reaction to emerging trial-external evidence and
+propose two novel adaptation methods that are consistent with the original design in the sense that the original design is only changed when the planning assumptions are modified.
+
+To this end, we first review assumed \mbox{conditional-,} observed \mbox{conditional-,} and predictive power
 from an estimation perspective.
-For sake of simplicity,
-we use a simple single-arm one-stage design to illustrate core
-characteristics.
 We then discuss the risks of constructing an
-`adaptive' two-stage design by na\"{\i}vely applying
+adaptive two-stage design by na\"{\i}vely applying
 a conditional-power-based sample size recalculation rule and the
@@
 The drawbacks of this na\"{\i}ve approach directly lead to the
-concept of optimal-two stage designs and
-we build on ideas presented in \cite{brannath2004,pilz2019}, and
-\cite{adoptrjss2020} to derive optimal designs for the situation studied in
-this manuscript.
+concept of optimal-two stage designs \cite{brannath2004,pilz2019,adoptrjss2020}.
 Without loss of generality, we focus on minimal expected sample size as optimality criterion.
-Optimal two-stage designs, by definition, cannot be made more
+These optimal two-stage designs, by definition, cannot be made more
 efficient by sample size recalculation.
@@
 This trial-external evidence might change during an ongoing study
-and may thus mandate a sample size recalculation.
+and may thus still mandate a sample size recalculation.
+
 In Section~\ref{sec:unplanned-adaptation},
-we discuss when an unplanned sample size recalculation is
-reasonable,
+we then discuss,
 how such a recalculation interacts with the concept of
-optimal two-stage designs,
-and the practical implementation.
+optimal two-stage designs.
+In particular, we demonstrate that commonly used methods
+are inconsistent in that they imply that that the original design needs to be modified for every possible interim outcome even if there is no change to the planning assumptions.
+We propose two novel methods that overcome this inconsistency.
 
@@
 we consider the case of a single-arm trial.
-All considerations can easily be extended to the more
-practically relevant two-arm case.
+All considerations can easily be extended to the two-arm case (see the application example in Section~\ref{2arm}).
 We further assume that the individually observed outcomes of the study participants $X_i, i = 1,\ldots,n$ are $iid$ and that their distribution has finite first moment $\theta$ and unit variance.
@@
 on the boundary of the null hypothesis, $Z_n\stackrel{\cdot}{\sim}\mathcal{N}(0, 1)$.
-
-Further assuming a maximal permissible type-I error rate of $\alpha=0.025$ (which we use for the remainder of the paper), the critical value for a single-stage fixed-sample-size design is the
+Further assuming a maximal permissible type~I error rate of $\alpha=0.025$ (which we use for the remainder of the paper), the critical value for a single-stage fixed-sample-size design is the
 $1-\alpha$ quantile of the standard normal distribution, i.e., approximately $c=1.96$.
 The test then rejects $\mathcal{H}_0$ if and only if $Z_n > c$ after the outcomes of $n$ subjects have been observed.
-The required size of the trial $n$, given a maximal allowable type-I error
+The required size of the trial $n$, given a maximal allowable type~I error
 rate, is usually determined by some form of restriction on
@@
 Approaches to defining such a power constraint under
-different assumptions were reviewed in~\cite{kunzmann2020}.
+different assumptions were reviewed in~Kunzmann~\etal\cite{kunzmann2020}.
 
-\section{Monitoring power} %%%%%%%%%%%%%%%%%%%%%%%
+\section{Previous work}
+
+\subsection{Monitoring power} %%%%%%%%%%%%%%%%%%%%%%%
 \label{sec:monitoring-power}
@@
 eventually rejecting the null hypothesis.
-Both the rejection probability conditional on the presence of an effect (conditional power) or the joint probability of rejection and presence of an effect (conditional probability of success) could be of interest.
-Here, we focus on the former due to the central role of
-power (and conditional power) in the planning of clinical trials.
-This effectively constitutes an estimation problem where the estimand is
-the \emph{conditional} probability of rejecting the null hypothesis given
-$Z_m=z_m$ and $\theta > 0$.
-Here the conditioning on a positive effect is crucial to justify the notion
-of `power' in contrast to the unconditional `probability of success'~\cite{kunzmann2020}.
-
 Since $Z_n$ and $Z_m$ are (asymptotically) jointly normal, the conditional distribution of $Z_n$ given $Z_m$ and $\theta$ is again normal and
@@
 \end{align}
-where $\tau:=m/n$ is the `information fraction' at the interim analysis.
-
+where $\tau:=m/n$ is the "information fraction" at the interim analysis (and the squared correlation between $Z_n$ and $Z_m$).
 The probability of rejecting the null hypothesis at the end of the trial
-as a function of $\theta$ is referred to as the `conditional power' in the literature.
-It is defined as
+as a function of $\theta$ is referred to as the conditional power in the literature.
+In the setting at hand, it is defined as
 \begin{align}
@@
 \end{align}
-Ignoring the implicit dependency on the time point of the interim analysis, $m$,
-conditional power is thus a function of the observed effect $\widehat{\theta}_m:=z_m/\sqrt{m}$
-(or equivalently the observed test statistic $z_m$), the critical value for the test decision~$c$,
-and the standardized effect size~$\theta$.
 Since $\theta$ is unknown,
-so is $\CP(z_m, c, \theta)$ and it thus cannot be
+so is $\CP(z_m, c, \theta)$ and it cannot be
 evaluated directly upon observing $Z_m=z_m$.
@@
 $\CP(z_m, c, \theta)$ can be estimated from observed data.
-The estimation perspective on `evaluating' conditional power is
+The estimation perspective on "evaluating" conditional power is
 less commonly taken in the literature but provides a consistent framework to
@@
 In a slight abuse of terminology,
-this quantity is often also referred to as `conditional power'.
+this quantity is often also referred to as "conditional power".
 To clearly distinguish it from $\CP(z_m, c, \theta)$ we denote
-it `assumed conditional power'
+it "assumed conditional power"
 \begin{align}
@@
 as a plug-in estimator for the unknown effect size~\cite{proschan1995}.
-This quantity is sometimes referred to as `observed conditional power'
+This quantity is sometimes referred to as "observed conditional power"
 in the literature and is defined as
@@
 prior probability density function evaluated at the parameter value $\theta$.
-Our definition of this so-called `predictive power'
+Our definition of this so-called "predictive power"
 \begin{align}
     \PP_\varphi(z_m,c) :&= \Pr_\varphi[\,Z_n > c \cond \Theta >0, Z_m = z_m\,] \\
-    &= \int_{0}^{\infty}\CP\big(z_m, c, \theta\big)\,\varphi\big(\theta\cond Z_m=z_m, \Theta\geq0\big)\operatorname{d}\theta
+    &= \int_{0}^{\infty}\CP\big(z_m, c, \theta\big)\,\varphi\big(\theta\cond Z_m=z_m, \Theta>0\big)\operatorname{d}\theta
 \end{align}
-differs slightly from the one proposed by~\cite{spiegelhalter1994}
+differs slightly from the one proposed by~Spiegelhalter~\etal\cite{spiegelhalter1994}
 in that we condition on a positive effect size $\Theta>0$.
-This is more consistent with the notion of frequentist power as discussed in~\cite{kunzmann2020}, although the difference is
+This is more consistent with the notion of expected power as discussed in~Kunzmann~\etal\cite{kunzmann2020}
+\begin{align}
+    \EP_\varphi(n, c) :&=
+    \Pr_\varphi\big[\,Z_n>c\,|\,\Theta>0\,\big]\\
+    &= \int \PP_\varphi(z_m,c)\ f_\varphi(z_m\mid \Theta>0) \,\operatorname{d}z_m
+\end{align}
+where $_\varphi(z_m\mid \Theta>0)$ is the probability density function of the predictive distribution of $Z_m\mid \Theta>0$.
+The difference is
 only of practical relevance when a substantial
@@
 computation of the posterior distribution under a normal likelihood.
-It is also the maximum entropy distribution on a compact interval given mean and standard deviation and thus `least informative'
-given a lower and upper boundaries on plausible effects as well as
+It is also the maximum entropy distribution on a compact interval given mean and standard deviation and thus "least informative"
+given a lower and upper boundary on plausible effects as well as
 the location and vagueness (variance) of the prior.
+The choice of the prior is a key consideration, but an extensive discussion is beyond the scope of this paper.
+We refer the reader to discussion in Spiegelhalter \textit{et al.}~\cite{spiegelhalter2004,rufibach_16}, and Hampson \textit{et al.}~\cite{hampson2021}, as well as the formal prior elicitation framework SHELF~\cite{kinnersley2013, shelf}, which is routinely used in practice by pharmaceutical companies~\cite{dallow2018}.
+
 Following~\cite{kunzmann2020} the required
-sample size $n$ is then determined by requiring
-a minimal expected power of $1-\beta=80\%$,
-where expected power is defined as
-\begin{align}
-    \EP_\varphi(n, c) :=
-    \Pr\big[\,Z_n>c\,|\,\Theta\geq0\,\big] = \int \Pr_\theta\big[\,Z_n>c\,\big]\, \varphi(\theta\,|\,\Theta\geq0) \,\operatorname{d}\theta \ .
-\end{align}
+sample size $n$ is determined by requiring
+a minimal expected power of $1-\beta=80\%$.
 Under these assumptions, the required sample size is
 $n = 79$.
-No point alternative is necessary to derive this sample size.
-We thus use the prior mean rounded to the first decimal point to
+Since no point alternative is necessary to derive this sample size, we use the prior mean rounded to the first decimal point to
 evaluate $\ACP$ and define $\theta_1:=0.4$.
@@
 The three estimators are depicted in Figure~\ref{fig:acp-ocp-pp-comparison-absolute} as functions
-of the observed interim outcome, where the interim time-point $m = 26$.
+of the observed interim outcome, where the interim time-point $m = 26$ (chosen heuristically as approximately 1/3 of the overall sample size).
 \begin{figure}
@@
     \caption{%
-        Properties of $\ACP,\OCP$, and $\PP$ as estimators of the unknown conditional power at $m=26$ an overall sample size $n=79$.
+        Properties of $\ACP,\OCP$, and $\PP$ as estimators of the unknown conditional power at $m=26$ and overall sample size $n=79$.
     }
@@
 $\PP$~on~${\Theta > 0}$.
-The practical implication of this can be seen
-by looking at the sampling distributions of the
-three estimators under different assumptions
-about the value of $\theta$ (see~Figure~\ref{fig:sampling-distributions}).
-Since $\OCP$ does not condition on $\Theta>0$, the sampling distribution is much more left-skewed for small effects than for the other two estimators.
+Since $\OCP$ does not condition on $\Theta>0$, the sampling distribution is much more left-skewed for small effects than for the other two estimators (see~Figure~\ref{fig:sampling-distributions}).
 Also, the negligence of the sampling variation of $\widehat{\theta}_m$, which $\OCP$ plugs in the expression for conditional power, leads to a larger variance for intermediate values of $\theta$ and the characteristic
@@
 see Figure~\ref{fig:acp-ocp-pp-as-estimators}.
-$\PP$ is the posterior expectation of conditional power with respect to the chosen prior and thus  minimizes the quadratic Bayes risk, i.e.~the MSE.
+$\PP$ is the posterior expectation of conditional power with respect to the chosen prior (conditional on a positive effect) and thus  minimizes the quadratic Bayes risk, i.e.~the average MSE weighted by $\varphi(\cdot\mid\Theta>0)$.
 This leads to relative good precision for parameter values with high \textit{a~priori} likelihood
@@
 $\PP$ is thus the natural choice for monitoring power during the course of a trial.
-Depending on the prior chose, it's properties can be either more similar to $\ACP$ (low prior variance) or $\OCP$ (high prior variance).
+Depending on the prior chosen, its properties can be either more similar to $\ACP$ (low prior variance) or $\OCP$ (high prior variance).
 
-\section{Na\"{\i}ve unplanned sample size adaptations}
+\subsection{Na\"{\i}ve unplanned sample size adaptations}
 \label{sec:naive-adaptations}
@@
 Typically, sample size recalculation is considered in the context of group-sequential designs at the pre-specified interim analysis \cite{proschan1995,brannath2004,bauer2016}.
-From a statistical perspective,
+From a statistical perspective, however,
 there is no reason why a sample size recalculation should not be
 performed for a study that initially was planned using a single-stage design.
-For the sake of simplicity the following considerations thus assume a simple
-single-stage design with fixed $n$ and $c$ as starting point and are
-restricted to a single interim analysis.
+For the sake of simplicity the following considerations are based on a single-stage design with fixed $n$ and $c$ as starting point.
+We restrict the discussion to a single unplanned interim analysis.
 All results can be generalized to the case of more complex starting designs and multiple interim analyses.
 
-A method for sample size recalculation that is often discussed in the literature is to derive a new sample size $n'$ and a new critical value $c'$ such that some estimate of conditional power exceeds a threshold $1-\beta_{cond.}$.
+A method for sample size recalculation that is often discussed in the literature is to derive a new sample size $n'$ and a new critical value $c'$ such that some estimate of conditional power exceeds a threshold $1-\widetilde{\beta}(z_m)$ that may or may not vary with the observed interim result.
 Different choices for the estimator of conditional power
-and the choice of $\beta_{cond.}$ have been discussed in the literature \cite{proschan1995,brannath2004,bauer2016}.
-Strict overall type-I error rate control can be maintained by invoking the conditional error principle \cite{muller2001,muller2004,brannath2012}, i.e., by limiting the maximal type-I error rate of the new design to the maximal conditional type-I error rate of the original design given the observed $z_m$.
-Often, trial protocols leave the exact choice of $1-\beta_{cond.}$ open since it can be chosen \textit{ad~hoc}
-without compromising strict type-I-error rate control.
-The basic concept of readjusting the sample size to achieve the desired conditional power is, however, a common approach, see for instance,~\cite{bhatt2013} and~\cite{mehta2009}.
+and the choice of $\widetilde{\beta}(z_m)$ have been discussed in the literature \cite{proschan1995,brannath2004,bauer2016}.
+Strict overall type~I error rate control can be maintained by invoking the conditional error principle \cite{muller2001,muller2004,brannath2012}, i.e., by limiting the maximal type~I error rate of the new design to the maximal conditional type~I error rate of the original design given the observed $z_m$.
+Often, trial protocols leave the exact choice of $1-\widetilde{\beta}(z_m)$ open since it can be chosen \textit{ad~hoc}
+without compromising strict type~I-error rate control.
+The basic concept of adjusting the sample size to achieve the desired conditional power is, however, a common approach, see for instance Bhatt~\etal~\cite{bhatt2013} and~Mehta~\etal\cite{mehta2009}.
 
@@
 we use predictive power as an estimator of conditional power.
-Furthermore, we set $\beta_{cond.}=\beta=0.2$.
+Furthermore, we set $\widetilde{\beta}(z_m)=\beta=0.2$.
 For given $Z_m = z_m$, the recalculation rule then corresponds to solving
@@
     \st             && \CP(m, n', z_m, c', 0) &\ \leq \ \CP_n(m, n, z_m, c, 0) \label{eq:ce-principle-naive-recalc} \\
-                    &&    \PP_\varphi(m, n', z_m, c') &\ \geq \ 1 - \beta_{cond.} \\
+                    &&    \PP_\varphi(m, n', z_m, c') &\ \geq \ 1 - \widetilde{\beta}(z_m) \\
                     &&                   n' &\ \geq \ n_{min} > m \label{eq:nmin-naive-recalc} \\
@@
 Constraint~\eqref{eq:ce-principle-naive-recalc} implements the conditional
-error principle by limiting the conditional (type-I) error rate under the new design to
-the conditional (type-I) error rate under the original design.
+error principle by limiting the conditional (type~I) error rate under the new design to
+the conditional (type~I) error rate under the original design.
 The minimal sample size constraint~\eqref{eq:nmin-naive-recalc} allows  $n_{min}>m$ for cases where a minimal sample size upon rejection of the null hypothesis is deemed necessary.
-This could e.g.~be of interest if a minimal precision of the final maximum likelihood estimator is sought.
 Note that the trial cannot stop at $n'=m$ without immediately accepting the
@@
 the trial is usually stopped early for futility declaring that the null hypothesis cannot be rejected.
-Alternatively, one could continue with the maximal sample size accepting
-the fact that the conditional power constraint is not met.
-This would, however, raise the question as to why a hard constraint on conditional power was imposed in the first place.
 
@@
 it yields two functions $n(z_m)$ and $c(z_m)$ as the point-wise solution of the problem
-which jointly define an `adaptive design' for any $z_m$.
-Here, `mandatory' means that it is decided \textit{a~priori} to always (for all $z_m$)
+which jointly define an adaptive design.
+Here, "mandatory" means that it is decided \textit{a~priori} to always (for all $z_m$)
 recalculate the
@@
 above specified way after a fixed number of outcomes $m$ has been observed.
-The term `adaptive design' is slightly misleading in this context.
+The term "adaptive design" is slightly misleading in this context.
 The design itself is pre-specified (and thus not adapted or changed)
 but only $n$ and $c$ are adaptive since they vary as functions of $z_m$.
-We would like to emphasize that such a detailed specification of the sample size recalculation rule is not common in practice since most protocols are focused on guaranteeing control of the maximal type-I-error rate control and maintaining maximal flexibility during the interim analysis.
-Still, if a sample-size reassessment was prescribed in the study protocol, there must have been some rationale as to the goal of that reassessment and we merely investigate the consequences for a design's unconditional properties when applying such a rationale in a mandatory way.
-The described approach should thus be seen as a hypothetical example highlighting the ineffectiveness that can arise from a na\"{\i}ve implementation of a recalculation based on a fixed threshold for
-conditional power.
 The corresponding sample size and critical value functions for $m=26$ are depicted in Figure~\ref{fig:binding-recalculation}.
@@
         with sample size $n'(z_m)$ and critical value $c'(z_m)$ functions defined by the conditional optimization problem \eqref{eq:naive-recalc}--\eqref{eq:nmax-naive-recalc} and
-        ${n=79}, {m=26}, {n_{max}=160}, {n_{min}=30}$ and $1-\beta_{cond.}=0.8$;
+        ${n=79}, {m=26}, {n_{max}=160}, {n_{min}=30}$ and $1-\widetilde{\beta}(z_m)=0.8$;
         the optimal design is the solution of \eqref{eq:optimal-design-start}--\eqref{eq:optimal-design-end}.
@@
 The mandatory application of the adaptation rule
-results in a two-stage design.
-The final sample size is thus a random variable $n(Z_m)$.
-This implies that the objective criterion of `minimal sample size' is no longer meaningful.
-Instead, one could use a weighted sum, $\E[\,n(Z_m)\,] + \eta \sqrt{\V[\,n(Z_m)\,]}$, of the expected value and standard deviation of the sample size as the objective criterion.
+results in a two-stage design and the final sample size is thus a random variable $n(Z_m)$.
+Optimality can thus no longer be defined in terms of the (random) sample size itself but only in terms of a functional of the distribution of $n(Z_m)$.
+For instance, one could use a weighted sum, $\E[\,n(Z_m)\,] + \eta \sqrt{\V[\,n(Z_m)\,]}$, of the expected value and standard deviation of the sample size as the objective criterion.
 The rationale for adding a penalty depending on the standard deviation of the sample size is that it might incur additional costs due to more
 complicated logistics (e.g., on demand production of additional drug doses).
-For the single-stage design this proposal always reduces to the fixed sample size $n$ since the standard deviation is 0.
-Whether or not the two-stage design is considered better then depends on the choice of $\eta$,
-i.e, the relative weight of the the standard deviation in the objective.
-In the particular situation considered here, the expected sample size is $49.5$ and its standard deviation is $29.7$.
+For the single-stage design this proposal always reduces to the fixed sample size $n$.
+Whether or not the two-stage design is considered better then depends on the choice of $\eta$ (the relative weight of the the standard deviation in the objective).
+In the particular situation considered here, the expected sample size is $48.3$ and its standard deviation is $30.1$.
 Since the original $n$ was $79$, the two-stage
-design would be considered `better'
-than the single-stage design for $\eta>0.99$.
+design would be considered better
+than the single-stage design for $\eta<1.02$.
 
-This comparison, however, ignores the fact that
-the original unconditional constraint of and expected power of more than $80\%$
+This comparison ignores the fact that
+the original unconditional constraint of an expected power of more than $80\%$
 is no longer fulfilled by the design with mandatory
 sample size adaptation.
-Both expected power ($72.1\%$ \emph{versus} $80.0\%$) and maximal type-I error rate ($2.1\%$ \emph{versus} $2.5\%$) are lower than for the original design.
+Both expected power ($66.9\%$ \emph{versus} $80.0\%$) and maximal type~I error rate ($1.8\%$ \emph{versus} $2.5\%$) are lower than for the original design.
 The difference in operating characteristics is mainly due to the fact
@@
 when the recalculation problem does not yield a $n'\leq n_{max}$.
-This effect is less pronounced when the same recalculation rule is applied
-to an initial two-stage group-sequential design
-which already defines binding early stopping boundaries
-for futility and efficacy.
-
 Similar to suggestions in the literature on binding futility stopping,
-one could modify the initial nominal $\alpha$ and the conditional power threshold $\beta_{cond.}$ until the recalculated design again fulfills the required
+one could modify the initial nominal $\alpha$ and the conditional power threshold $\widetilde{\beta}(z_m)$ until the recalculated design again fulfills the required
 operating characteristics \cite{brannath2004}.
 This approach to define a fully pre-specified two-stage design is, however, unnecessarily complicated in that it
-uses methods like the conditional error principle or a combination function, which are originally intended for \emph{unplanned}
-design adaptations.
+uses methods which are originally intended for \emph{unplanned}
+design adaptations and that
+they still implicitly depend on a design that is never actually realized via the
+conditional error constraint~\eqref{eq:ce-principle-naive-recalc}.
 
-\section{Optimal pre-planned sample size adaptations}
+\subsection{Optimal pre-planned sample size adaptations}
 \label{sec:optimal-adaptation}
-
-Evidently, the na\"{\i}ve application of the predictive power
-adaptation rule outlined earlier is needlessly conservative.
-Fixing the operating characteristics by tuning the nominal $\alpha$ and
-$\beta_{cond.}$ does not address the deeper issue that the sample size and
-critical value functions are still derived based on a merely heuristic recalculation criterion and that
-they still implicitly depend on an original design that is never actually realized via the
-conditional error constraint~\eqref{eq:ce-principle-naive-recalc}.
 
@@
 can be optimized directly.
-For the sample size function $n(z_m)$ \cite{brannath2004} discussed this approach by fitting a polynomial of degree 4 but they did not optimize
+For the sample size function $n(z_m)$ Brannath~\textit{et al.}~\cite{brannath2004} discussed this approach by fitting a polynomial of degree 4 but they did not optimize
 any of the other parameters.
-In a more general form, \cite{pilz2019} approached the problem using
+Pilz~\textit{et~al.}~\cite{pilz2019} approached the problem in a more general form using
 variational methods.
 An \texttt{R} implementation that uses cubic splines for both the $n$ and
-$c$ functions is available via the package~\textit{adoptr}~\cite{adoptrjss2020}.
+$c$ functions is available via the package~\texttt{adoptr}~\cite{adoptrjss2020}.
 
 In the following we restrict the considerations to the case
-where only expected sample size is of interest and the variation is
+where only expected sample size is of interest and the variability of the sample size is
 ignored, i.e.~$\eta=0$.
@@
 and can be derived numerically using the \texttt{R} package \texttt{adoptr}~\cite{adoptrjss2020}.
+In the following, "optimal" always refers to optimal with respect to expected sample size in the above sense.
 Here, $\E_\varphi[\,n(Z_m)\,]$ is the expected sample size under the prior $\varphi$ and expected power needs to be redefined to account for the
@@
     \EP_\varphi\big(n(\cdot), c(\cdot)\big) :=
-    \Pr_\varphi\big[\,Z_{n(Z_m)}>c(Z_m)\,|\,\Theta\geq0\,\big] \ .
+    \Pr_\varphi\big[\,Z_{n(Z_m)}>c(Z_m)\,|\,\Theta>0\,\big] \ .
 \end{align}
-Figure~\ref{fig:binding-recalculation} shows the sample size-, critical value-, and the predictive power function in the example situation discussed earlier together with the single-stage design
+Figure~\ref{fig:binding-recalculation} shows the sample size-, critical value-, and the predictive power function in the situation discussed earlier together with the single-stage design
 and the na\"{\i}ve adaptation based on predictive power discussed in
 Section~\ref{sec:naive-adaptations}.
-The optimal two-stage design complies with both the predictive power and the maximal type-I error rate constraints and is thus comparable with the original single-stage design in terms of sample size.
+The optimal two-stage design complies with both the predictive power and the maximal type~I error rate constraints and is thus comparable with the original single-stage design in terms of sample size.
 The expected sample size is much lower ($56.4$ \emph{versus} $79$) at the cost of a non-zero standard deviation of the expected sample size ($28.5$ \emph{versus} $0$).
 Since it is entirely pre-specified,
-the optimal two-stage design does not need to fall back to the conditional error principle to control the maximal type-I error rate.
+the optimal two-stage design does not need to fall back to the conditional error principle to control the maximal type~I error rate.
 Instead, it achieves the desired design characteristics in an optimal way since they are incorporated as constraints to problem \eqref{eq:optimal-design-start}--\eqref{eq:optimal-design-end}.
-The optimal design fully exhausts the allowable maximal type-I error rate
+The optimal design fully exhausts the allowable maximal type~I error rate
 and complies with the expected power constraint.
@@
 shape has a mode close to the early-futility boundary.
-A recalculation based on exceeding a fixed lower threshold for
+A~recalculation based on exceeding a fixed lower threshold for
 some estimator of conditional power always results in a convex
-sample size function and can thus never be fully optimal irrespective of how the nominal values of $\alpha$ and $\beta_{cond.}$
+sample size function and can thus never be fully optimal irrespective of how the nominal values of $\alpha$ and $\widetilde{\beta}(z_m)$
 are chosen.
+For a more detailed discussion of this phenomenon see~e.g.~\citep{pilz2019,brannath2004}.
 
 The fact that the optimal two-stage design exhibits a monotonously increasing
-predictive power rather than a constant predictive power of, say, 80\%
+predictive power rather than a constant predictive power
 indicates that any recalculation rule keeping predictive power
@@
 of the resulting design.
-It may thus appear as if any `recalculation' of an optimal design's sample size using methods
-for unplanned interim analyses was not only
-unnecessary but even counterproductive.
-This is, however, only the case if the initial design was optimal and
-the planning assumptions still hold.
-
-\section{Consistent unplanned recalculation}
-\label{sec:unplanned-adaptation}
-
-The preceding section made the case that the direct optimization of all design parameters is superior to the mandatory application of methods for unplanned design adaptations during the planning phase of a trial.
-By definition,
-no optimal design ever needs to be changed as a reaction to trial internal data during the course of the study.
+This is inconsistent with the original optimization objective -
+by definition,
+no optimal design ever needs to be changed as a reaction to trial internal data alone during the course of the study.
+The direct optimization of all design parameters is superior to the mandatory application of methods for unplanned design adaptations during the planning phase of a trial.
 However, there are still situations in which an unplanned sample size recalculation is warranted.
@@
 
-In the following, we discuss two approaches of reacting to such trial-external events in a consistent way
-if the original design minimizes expected sample size.
-Here, `consistent' means that the original optimal design is invariant under
-the recalculation rule unless the planning assumptions (prior) are changed or
-the time-point of the interim analysis is different from the optimal one.
-The considerations are restricted to cases where the result of an adaptation is a final sample size and no further interim analyses are planned for the remainder of the trial.
+\section{Consistent unplanned sample size adaptations}
+\label{sec:unplanned-adaptation}
+
+In the following, we discuss two approaches of reacting to
+trial-external events during an unplanned interim analysis in a consistent way.
+Here, "consistent" means that the original design is invariant under
+the sample size recalculation rule for all potential interim
+results under the original design unless the planning assumptions or
+the time-point of the interim analysis are changed.
+An example for such an inconsistent sample size recalculation rule was given in Section~\ref{sec:naive-adaptations}:~The design optimizing expected sample size is not invariant under a rule that adapts the sample size and the critical value to match a predictive power of exactly $80\%$.
+The same holds for a simpler group-sequential design minimizing expected sample size
+since the predictive power of a group-sequential trial cannot be constant on the continuation area.
+The following considerations are restricted to cases where the result of an adaptation is a final sample size and no further interim analyses are planned for the remainder of the trial.
 This means that the unplanned adaptation replaces the pre-planned interim analysis although it might occur at a different point in time.
-All methods can be generalized to an unplanned interim analysis in the
-first stage with a subsequent second interim analysis.
-This might be necessary when the unplanned interim analysis has to
-be conducted shortly after the start of a trial.
 
-To obtain a consistent sample size recalculation rule,
-the conditional error principle can be applied not only to the
-maximal type-I error rate but also to the (average) type-II error rate.
-Let $m'$ be the sample size of the unplanned interim analysis and
-$\psi$ be the probability density function of the revised prior.
-We propose to recalculate the new sample size, $n'$ and
-the new critical value $c'$ as the solution of the point-wise optimization problem
-\begin{align}
-    \argmin{n',c'} && n' &\\
-     \st && \CP(m', n', z_{m'}, c', 0) &\leq \Pr_0[\,Z_{n(Z_m)} > c(Z_m) \cond Z_{m'} = z_{m'}\,] \label{eq:conditional-error-recalc}\\
-     && \PP_\psi(m', n', z_{m'}, c') &\geq \nonumber \\
-     && & \hspace*{-1cm}\Pr_\varphi[\,Z_{n(Z_m)} > c(Z_m) \cond \Theta >0, Z_{m'} = z_{m'}\,] \label{eq:pp-recalc} \ .
-\end{align}
-Here, constraint~\eqref{eq:pp-recalc} means that the modified design under the new prior has at least as much predictive power as the original design under the original prior.
-Constraint~\eqref{eq:conditional-error-recalc} implements the conditional error principle and thus ensures strict type-I error rate control of the
-procedure.
-The crucial difference to the procedure discussed in Section~\ref{sec:unplanned-adaptation} is the fact that the lower boundary on predictive power varies with $z_m$.
-
-In cases where the interim analysis is conducted at the original $m$,
-the problem simplifies to
+We consider minimal expected sample size as original objective criterion.
+Since the sample size after the interim analysis is fixed,
+the corresponding conditional objective criterion is the minimization of the second-stage sample size.
+As in Section~\ref{sec:naive-adaptations}, the overall type~I error rate is controlled via a
+constraint on the conditional type~I error rate.
+W.l.o.g.,~we only consider cases where the original design would not stop early for futility or efficacy.
+Let $\psi$ be the probability density function of the revised prior.
+We propose to recalculate the new sample size $n'$ and
+the new critical value $c'$ as the solution of the point-wise (conditional on $Z_{m'} = z_{m'}$) optimization problem
 \begin{align}
@@
     && \CP(m, n', z_{m}, c', 0) &\leq \CP(m, n(z_m), z_{m}, c(z_m), 0) \\
-    && \PP_\psi(m, n', z_{m}, c') &\geq \PP_\varphi(m, , n(z_m), z_{m}, c(z_m))  \label{eq:pp-original-interim} \ .
+    && \PP_\psi(m, n', z_{m}, c') &\geq 1 - \widetilde{\beta}(z_m) \ . \label{eq:objective-original-interim-end}
 \end{align}
-Since $\CP$ is monotone in $n'$, the solution is already uniquely
+Since $\CP$ is monotone in $n'$, the solution is uniquely
 defined by the constraints, if it exists.
-It follows directly that, for $m'=m, \psi=\varphi$,
-the original design ${n'=n(z_{m})}$, ${c'=c(z_m)}$ fulfills both constraints with equality.
+The optimization thus reduces to finding the root of
+\begin{align}
+    \solve{n',c'} \qquad
+    & \CP(m, n', z_{m}, c', 0) &&= \CP(m, n(z_m), z_{m}, c(z_m), 0) \label{eq:root-cp}\\
+    & \PP_\psi(m, n', z_{m}, c') &&= 1 - \widetilde{\beta}(z_m) \ . \label{eq:root-pp}
+\end{align}
+The central problem is translating the unconditional (expected) power constraint of the
+original design problem into a conditional one for predictive power - i.e.~to pick a value for $\widetilde{\beta}(z_m)$ - in a way that ensures consistency of the recalculated sample size and critical value with the original design if neither the interim timing nor the planning assumptions change.
+As discussed in Section~\ref{sec:naive-adaptations}, a constant value for $\widetilde{\beta}(z_m)$ does not, in general, lead to a consistent recalculation rule.
+Below, we propose two consistent recalculation rules,
+both are based on an application of the conditional error principle to the (expected) type~II error rate.
+The first rule does not require knowledge of the underlying optimization problem for the original design.
+This approach is easy to implement but does not maintain the unconditional power properties of the design, which is addressed by the second proposal.
+
+\subsection{"Not-worse" approach}
+\label{sec:not-worse}
+
+A sensible heuristic criterion to chose $\widetilde{\beta}(z_m)$ is to require that the modified design under the new prior $\psi$ should have at least as much predictive power as the original design under the original prior $\varphi$,
+i.e.~$\widetilde{\beta}(z_m) = {1 - \PP_\varphi(m, n(z_m), z_{m}, c(z_m))}$.
+Consequently, we term this choice of $\widetilde{\beta}(z_m)$ the "not-worse" approach.
+It follows that, for $\psi=\varphi$,
+the original design's $n(z_m)$ and $c(z_m)$ are a root of \eqref{eq:root-cp}-\eqref{eq:root-pp} and hence ${n'=n(z_{m})}$, ${c'=c(z_m)}$.
 The original optimal design is thus indeed invariant under the
-proposed recalculation rule if neither the planning prior not the interim time-point are changed.
+proposed recalculation rule if neither the planning prior nor the interim time-point are changed.
+However, the
+expected power of the procedure resulting from always switching to
+a new prior $\psi\neq\varphi$ will be different from the originally targeted $1-\beta$ since $1-\widetilde{\beta}(z_m)$ does not integrate to $1 - \beta$ under~$\psi\neq\varphi$.
+To see this, let $f_\psi(z_m \mid \Theta>0)$ be the probability density function of the interim result $Z_m$ under $\theta\sim\psi$ conditional on $\Theta>0$, then
+{\small
+\begin{align}
+    \EP_\psi\big(n'(\cdot), c'(\cdot)\big) &=
+    \Pr_\varphi\big[\,Z_{n'(Z_m)}>c'(Z_m)\,|\,\Theta>0\,\big]  \\
+    &= \int \PP_\psi(m, n'(z_m), z_{m}, c'(z_m)) \ f_\psi(z_m \mid \Theta > 0) \operatorname{d} z_m \\
+    &= \int \underbrace{\big(1 - \widetilde{\beta}(z_{m})\big)}_{\PP_\varphi(m, n(z_m), z_{m}, c(z_m))} \ f_\psi(z_m \mid \Theta > 0) \operatorname{d} z_m \\
+    &\neq \int \PP_\varphi(m, n(z_m), z_{m}, c(z_m)) \ f_\varphi(z_m \mid \Theta > 0) \operatorname{d} z_m \\
+    &= \EP_\varphi\big(n(\cdot), c(\cdot)\big) = 1 - \beta \ .
+\end{align}}
 
-Problem \eqref{eq:conditional-error-recalc}--\eqref{eq:pp-recalc} is in no way the only conceivable
-means of deriving a consistently recalculated sample size.
-For the sake of simplicity, we consider the case of $m'=m$ (i.e.~problem \eqref{eq:objective-original-interim}--\eqref{eq:pp-original-interim}).
-Depending on how much the prior changes,
-the required sample size adjustment to meet the predictive power constraint can be quite drastic.
-In practice, it might thus be preferable to relax the predictive power constraint in a principled way.
-This can be done via the Lagrangian mechanism.
-Let
+\subsection{"Reoptimization" approach}
+\label{sec:reoptimize}
+
+To additionally maintain the unconditional (expected) power during an unplanned sample size recalculation,
+$1-\widetilde{\beta}(z_m)$ needs to integrate to $1-\beta$ under the modified prior $\psi\neq\varphi$.
+We propose to achieve this with a two-step procedure.
+Firstly, a modified version of the original optimization problem \eqref{eq:optimal-design-start}-\eqref{eq:optimal-design-end} is solved where the original prior~$\varphi$ is replaced with the updated prior~$\psi$,
+and under the additional conditional type~I error rate constraint dictated by the conditional error principle, i.e.
+{\small\begin{align}
+    \argmin{n''(\cdot), c''(\cdot)}
+        &&                                   \E_\psi[\,n''(Z_m)\,] & \\
+    \st &&                 \Pr_{0}[\,Z_{n''(Z_m)} > c''(Z_m)\,] & \leq \alpha \\
+        &&     \EP_\psi\big(n''(\cdot), c''(\cdot)\big) & \geq 1 - \beta \\
+        && \CP(m, n''(z_{m}), z_{m}, c''(z_{m}), 0) &\leq \CP(m, n(z_m), z_{m}, c(z_m), 0) \label{eq:cp-reoptimize}
+\end{align}}
+
+\noindent One could use $n''(z_m)$ as recalculated sample size and the corresponding critical value~$c''(z_m)$ directly.
+However, there may be cases where the conditional type~I error rate constraint \eqref{eq:cp-reoptimize} is not binding leading to an unnecessarily low conditional error and thus a needlessly large sample size.
+Instead, as the second step, we propose to plug in  $1-\widetilde{\beta}(z_m)=\PP_\psi(m, n''(z_m), z_{m}, c''(z_m))$ in \eqref{eq:root-cp}-\eqref{eq:root-pp} and solve for $n'$ and $c'$.
+Since,
+{\small
 \begin{align}
-    f(n', c') :&= n' \\
-    g(n', c') :&= \CP(m, n', z_{m}, c', 0) - \CP(m, n(z_m), z_{m}, c(z_m), 0) \\
-    h_\psi(n', c') :&= \PP_\varphi(m, , n(z_m), z_{m}, c(z_m)) - \PP_\psi(m, n', z_{m}, c')
-\end{align}
-As discussed earlier, the original designs $n$ and $c$ trivially are the unique solution of
-\begin{align}
-    \argmin{n',c'} && f(n',c') & \\
-    \st
-    && g(n',c') &\leq 0 \\
-    && h_\varphi(n',c') &\leq 0 \ .
-\end{align}
-This problem is equivalent to solving the unconstrained problem
-\begin{align}
-    \argmin{n',c'} f(n',c') - \lambda_g g(n',c') - \lambda_h h_\varphi(n',c')
-\end{align}
-for suitable choice of the Lagrange multipliers $\lambda_g$ and $\lambda_h$.
-Since the original $n, c$ are optimal,
-$\lambda_g$ and $\lambda_h$ can be obtained directly as solution of
-\begin{align}
-    0 &= \nabla f(n, c) - \lambda_g \nabla g(n, c) - \lambda_h \nabla h_\varphi(n, c) \\
-   \Leftrightarrow (1, 0)^\top &= \big(\nabla g(n, c), \nabla h_\varphi(n, c)\big) \cdot (\lambda_g, \lambda_h)^\top \\
-   \Leftrightarrow (\lambda_g, \lambda_h)^\top &= \big(\nabla g(n, c), \nabla h_\varphi(n, c)\big)^{-1} \cdot (1, 0)^\top\ .
-\end{align}
-Here, $\nabla$ denotes the gradient as column vector.
-The Lagrange multiplier corresponding to the predictive power constraint, $\lambda_h$, encodes the trade-off between predictive power and sample size of the original,
-optimal design at $z_m$.
+    1-\beta &\leq\EP_\psi\big(n''(\cdot), c''(\cdot)\big) \\
+    &= \int \underbrace{\PP_\psi(m, n''(z_m), z_{m}, c''(z_m))}_{1-\widetilde{\beta}(z_m)} \ f_\psi(z_m \mid \Theta > 0) \operatorname{d} z_m \ ,
+\end{align}}
 
-A recalculated sample size under a different prior $\psi$ can then be obtained by
-keeping $\lambda_h$ fixed and minimizing
-\begin{align}
-    \argmin{n',c'} && n'&-\lambda_h\, h_\psi(n',c') \\
-     \st && g(n', c') &\leq 0 \ .
-\end{align}
-This effectively relaxes the predictive power constraint by allowing the same trade-off
-between sample size and predictive power that was implied by the original optimal design.
-The conditional type-I error rate constraint is not relaxed to comply with the
-conditional error principle.
-This approach is closely related to the formulation of the optimal-design problem given in~\cite{jennison2015} and the utility perspective on sample
-size derivation presented in~\cite{kunzmann2020}.
-In contrast to the situation analyzed by \cite{jennison2015}, however,
-$\lambda_h$ does depend on $z_m$.
-This is due to the fact that expected sample size is formed with respect to the unconditional
-prior while expected power uses the prior conditional on $\Theta>0$.
+\noindent the \emph{unconditional} expected power of the recalculation procedure maintains the original value of $1-\beta$.
+Note that this only holds true if the choice of~$\psi$ is stochastically independent of $Z_m$, i.e.~if the change of the planning assumptions is not driven by the observed interim data.
 
-We illustrate both approaches by example of the optimal two-stage design that minimizes the objective
-$\E_\varphi[\,n(Z_m)\,]$ subject to a maximal type-I error rate of $0.025$ and a minimal expected power of $0.8$, see~Figure~\ref{fig:binding-recalculation}.
-This design has a pre-planned interim analysis after $m=35$ individuals.
-Now assume that
-the prior $\varphi$ needs to be revised due to new trial-external
-information at the pre-planned interim analysis.
-We consider members of the class of shifted truncated Normal distributions
-\begin{align}
-    \psi_{\mu}(\theta) := \frac{\displaystyle \boldsymbol{1}_{[-0.5, 1]}(\theta)\,\phi\left(\frac{\theta - \mu}{0.2}\right)}{\displaystyle \Phi\left(\frac{1 - \mu}{0.2}\right) - \Phi\left(\frac{-0.5 - \mu}{0.2}\right)}
-\end{align}
-as revised priors, i.e. the original prior~$\varphi=\psi_{0.4}$.
-Figure~\ref{fig:recalculation-old-m} shows the behavior of the recalculated sample size and critical value in this scenario for observed effects of $z_m/\sqrt{m}=0.25, 0.3, 0.35$ and varying mean of the revised prior if the originally planned interim time-point is maintained.
+\subsection{Application Example}
+\label{2arm}
+
+In practice, group-sequential designs are more common than optimal two-stage designs.
+Since the efficiency gains of the latter over the simpler group-sequential designs
+are often practically irrelevant this is a sensible approximation~\citep{wassmer2016}.
+Sample size calculations are still commonly based on point alternatives.
+This is equivalent to using a point prior on the alternative value.
+All previous considerations can thus be transferred to a setting with a fixed
+point alternative.
+
+For this example, we consider the two-stage, two-arm group-sequential design with binding early stopping for futility boundary at an observed effect of $0$ that minimizes the expected sample size under the point alternative of a mean difference of $\theta_1=0.4$ under a one-sided type~I error rate constraint of $\alpha \leq 0.025$ and  an expected power of at least $0.8$.
+Since the prior only has mass at $\theta_1=0.4$ this corresponds to a classical power constraint at $\theta_1=0.4$.
+We computed the optimal group-sequential design by direct minimization over the stage-one and two sample sizes and the stage-two critical value using the \texttt{R}~package \texttt{nloptr}~\cite{nloptr}.
+
+We first consider a scenario where the alternative is modified during the interim analysis to either $\theta_1=0.35$ or $\theta_1=0.45$ \emph{independently} of the
+observed interim data.
+The conditional and unconditional properties of three adaptation procedures together with those of the original design are shown in Figure~\ref{fig:application-example-1}.
+Firstly, "PP >= 0.8" corresponds to a recalculation based on a fixed threshold for the predictive power of $0.8$.
+As discussed earlier, this adaptation rule is not consistent in that the original design is not invariant under recalculation (see Section~\ref{sec:naive-adaptations}).
+Secondly, the "not-worse" approach discussed in Section~\ref{sec:not-worse} instead matched the predictive power of the original design under the old prior.
+Thirdly, the "reoptimize" approach discussed in Section~\ref{sec:reoptimize} matches the predictive power of design that is optimal under the new alternative.
+For all recalculations, the final adapted sample size was restricted to the range of $120$ to $400$.
 \begin{figure}
     \centering
-    \begin{subfigure}[b]{\textwidth}
+    \includegraphics[width=\textwidth]{figures/application-example}
+    \caption{%
+        Left: Sample size (both arms) and predictive power for response-independent adjustment of the alternative; predictive power is evaluated at the respective changed point alternative.
+        Right: unconditional (expected) power and expected sample size for the three scenarios when adapting to the respective true $\theta$ (for $\theta=0.4$, the two consistent rules overlap with the original design).
+    }
+    \label{fig:application-example-1}
+\end{figure}
+Clearly, "PP >= 0.8" is not only inconsistent but also the rule that leads to the most extreme adaptation of the sample size.
+The "not-worse" approach leads to a level-shift in the sample size with moderate
+increases or decrease of the stage-two sample size.
+The "reoptimize" approach leads to more moderate increases in sample size
+as compared o the "PP >= 0.8" approach when the alternative is lowered and to substantial sample size reductions for an increased effect size.
+From an unconditional perspective, only the "reoptimize" approach maintains the target expected power of $0.8$.
+
+A conceptual drawback of point priors is that they are invariant under
+formal Bayesian updating.
+This means that the posterior distribution after observing the interim results
+is always the same point measure as the initial point prior
+and is merely a consequence of claiming perfect knowledge of the effect size during planning.
+In practice,
+investigators might still be inclined to modify the alternative at least heuristically during the interim analysis.
+We thus consider an alternative situation based on the same optimal group-sequential design discussed above.
+Assume that the trial team wants to revise the strong assumption of a point prior on the alternative of a mean difference of $0.4$ during the interim analysis using the observed interim data.
+For instance this could be done heuristically by "updating" the alternative to the average of the observed effect and the originally assumed point alternative of $0.4$.
+The resulting conditional and unconditional properties of the proposed recalculation rules are given in Figure~\ref{fig:application-examplel-response-adaptive}.
+\begin{figure}
     \centering
-        \includegraphics[width=.8\textwidth]{figures/recalculation-type-II-original-m}
+    \includegraphics[width=\textwidth]{figures/application-example-response-adaptive.pdf}
     \caption{%
-            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect.
+        Left: Sample size (both arms) and predictive power for response-adaptive modification of the alternative; predictive power is evaluated at the respective changed point alternative.
+        Right: unconditional (expected) power and expected sample size for three scenarios.
     }
-        \label{fig:recalculation-old-m}
-    \end{subfigure}
-    \begin{subfigure}[b]{\textwidth}
-        \centering
-        \includegraphics[width=.8\textwidth]{figures/recalculation-type-II-earlier}
-        \caption{%
-            Earlier-than-planned interim analysis at an observed effect of $0.3$ and varying interim time-point.
-        }
-        \label{fig:recalculation-earlier}
-    \end{subfigure}
-    \begin{subfigure}[b]{\textwidth}
-        \centering
-        \includegraphics[width=.8\textwidth]{figures/recalculation-lambda-approach-original-m}
-        \caption{%
-            Interim analysis at the originally planned interim sample size of $m=35$ for varying observed effect and the $\lambda$ approach.
-        }
-        \label{fig:recalculation-lambda}
-    \end{subfigure}
-    \caption{%
-        Comparison of consistent unplanned recalculation schemes.
-    }
+    \label{fig:application-examplel-response-adaptive}
 \end{figure}
-Clearly, the original design is indeed invariant if the prior remains unchanged.
-The sample size is increased if the new prior is more conservative
-than $\varphi$ and \textit{vice~versa}.
+In this alternative scenario, the adapted sample size under all three rules is more variable and quickly reaches the maximal value of $400$.
+In this scenario, the sample size is rarely reduced since the original group-sequential design employs an aggressive early-efficacy boundary of $\approx 0.4$ on the scale of the observed treatment effect.
+Hence, the heuristically modified alternative almost never exceeds the original value of $\theta_1 = 0.4$.
+Since the adaptation is now response-adaptive, the "reoptimize" approach does no longer maintain a power of $1-\beta$.
 
-Of course, it could also be necessary to conduct the interim analysis earlier or later than originally anticipated, at $m'\neq m$.
-In this case, the design is no longer invariant since the earlier
-or later time point already implies a deviation from the
-original optimal design even if the prior remains the same.
-Figure~\ref{fig:recalculation-earlier} shows the resulting adaptations
-for earlier-than-planned time points of the interim analysis
-while the observed effect is kept fixed at $0.3$.
-The formula for the conditional error in this situation is given in the Appendix.
-
-Figure~\ref{fig:recalculation-lambda} shows the recalculated sample size
-for $m'=m=35$ and the $\lambda$-approach.
-The original design is again (approximately) invariant under recalculation.
-Compared to the fixed conditional type-II error approach, predictive power
-under the $\lambda$-approach is more flexible and might drop below the predictive power under the original design, see Figure~\ref{fig:recalculation-lambda}, right panel.
-This flexibility in terms of predictive power leads to a much smaller deviation from the original design's sample size.
-The $\lambda$-approach transitions smoothly from
-an increase of the original sample size due to a slightly more conservative prior to a decrease under very conservative priors.
-This behavior leads to less extreme adjustments and is thus
-more realistic in practice.
-After all, a slight loss in (predictive) power as compared to the
-original design might be tolerable if the sample size can be kept at
-a moderate level in exchange.
-Of course, it is still possible to stop the trial early for futility
-at any point without compromising type one error rate control
-if the predictive power is considered too low to warrant a continuation.
+These two example situations show that a consistent sample size adaptation can make sense as reaction to new trial-external evidence.
+For the relevant case of reacting to new trial-external evidence,
+it is possible to maintain the unconditional power properties of the original design.
+If, however, the change of alternative depends on the observed interim data,
+it is unclear how the unconditional power may be preserved.
 
@@
 is a function of the unknown treatment effect.
-An often overlooked aspect of this estimation problem is that it is
-conditional on a positive treatment effect since the notion of power is conditional on there being an non-null effect~\cite{kunzmann2020}.
-In many practical situations where the prior mass is concentrated on
-non-null effects the effect of conditioning is negligible but its importance increases with the vagueness of a the prior.
-This lack of conditioning on a positive treatment effect is one of the reasons observed conditional power performs rather poorly as an estimator of the unknown conditional power.
+This estimation problem can be formulated both conditional or unconditional on a positive treatment effect.
+In many practical situations where the prior mass is concentrated on non-null effects the consequences of conditioning are negligible but their importance increases with the vagueness of a the prior.
+The lack of conditioning on a positive treatment effect explains why observed conditional power performs rather poorly as an estimator of the unknown conditional power for a wide range of effect sizes.
 Assumed conditional power for $\theta_1>0$ does reflect this conditionality.
@@
 Altering a simple starting design with a seemingly intuitive recalculation rule
-can actually lead to overall \emph{less effective} trials.
+can lead to overall \emph{less effective} trials.
 Techniques originally intended for \emph{unplanned} adaptations of trials
@@
 By definition, no trial-internal event can then justify a sample size recalculation.
-Only trial external events, such as a change in the objective criterion,
+Only trial-external events, such as a change in the objective criterion,
 the emergence of new trial-external evidence,
@@
 may require a reassessment of the sample size of an optimal design.
-A generic recalculation rule might lead to the paradoxical situation that even an optimal
+
+A~generic recalculation rule might lead to the paradoxical situation that an optimal
 starting design is always modified during the interim analysis
@@
 if the planning assumptions remain unchanged.
-For instance, this minimal consistency property is not fulfilled when recalculating a design's
-sample size based on a fixed threshold for its minimal conditional power
-since designs optimizing expected sample size tend to have variable conditional power.
+This minimal consistency property is often not fulfilled when recalculating a design's
+sample size based on a fixed threshold for its minimal conditional power.
 
-We propose two consistent approaches to adjusting an ongoing optimal design to
+We proposed two consistent approaches to adjusting an ongoing optimal design to
 newly emerging trial-external data.
-The first method applies the conditional error principle to the (average) type-II error rate in a
-similar way to its use in controlling the maximal type-I error rate.
+Both methods can be extended to deviations from the planned interim time-point.
+The first method ("not-worse") applies the conditional error principle to the (average) type~II error rate in a similar way to its use in controlling the maximal type~I error rate.
 The method is easy to implement and consistent in the above defined way.
-We would like to stress that this is by no means the only way of
-conducting a consistent sample size recalculation.
-The alternative $\lambda$-approach allows a trade-off between predictive power
-and sample size leading to smaller sample size deviations from the original design.
-This comes at the risk of potentially loosing predictive power.
-Both methods differ by the quantity that is held fixed during the recalculation.
-For the first method, the predictive power under the old design and the old prior
-is maintained for the new design under the new prior whereas the $\lambda$-approach
-only keeps the trade-off between predictive power and sample size fixed but allows the
-actual predictive power to deviate form the original design.
+However, it does not maintain the same unconditional power as the original design,
+not even under data independent modifications of the planning assumptions.
+The second method addresses this issue by adapting the sample size and critical value such that the predictive power matches the predictive power of a design that would have been optimal, had the new prior been known during the planning stage.
+As long as the prior is adapted data-independently, this approach does indeed maintain the same unconditional power properties.
 
-These considerations show how absolutely crucial the initial planning stage of a trial is.
+These considerations show how crucial the initial planning stage of a trial is.
 Adaptive methods should not be taken as an excuse to start with a
@@
 Ideally, all uncertainty is quantified during the planning phase to
-the best possible extend and integrated in the initial design via a
+the best possible extent and integrated in the initial design via a
 planning prior.
+A modification of the design is then only necessary and warranted
+when this planning prior changes.
 This does not mean that a complex design is always the best choice.
@@
 perform better than more complex alternatives.
-Also,
-optimal-two stage designs require optimizing over function-spaces to find the optimal sample size and critical value functions.
+Generic optimal-two stage designs require optimizing over function-spaces to find the optimal sample size and critical value functions.
 Obtaining a stable solution might thus be hard in practice and
@@
 Still, the same principle considerations apply:
-An optimal (group-sequential) design only needs to be revised if either the objective function or the underlying planning assumptions, i.e., the prior on the effect size, changes.
-
-Adaptations of initially sub-optimal designs are another potential
-application of the new recalculation scheme.
-Clearly, the single-stage design with $n=79$ is not minimizing
-\emph{expected} sample size.
-To switch to a near-optimal adaptive design mid-trial,
-one could derive the optimal two-stage design given the current best prior \textit{post-hoc} and
-then use the $\lambda$-approach to recalculate the initial design while controlling the maximal
-conditional type-I error rate in the usual way.
-The resulting design will not be optimal from an unconditional perspective since the conditional error function of the initial, sub-optimal design has to be respected but the power-sample size trade-off during the interim analysis
-will be similar to the optimal design due to the choice of $\lambda_h$.
+An optimal (group-sequential) design only needs to be revised if either the objective function or the underlying planning assumptions, i.e.~the prior on the effect size, changes.
 
 A limitation of the proposed recalculation methods is the fact that they
-rely on the conditional error principle for strict type-I-error rate control.
+rely on the conditional error principle for strict type~I-error rate control.
 The conditional error principle can be difficult to extend to cases with nuisance
 parameters \cite{gutjahr2011}.
-In practice, a simple plug-in approach similar to the approach in blinded sample size reassessment might be viable but have not yet been investigated more thoroughly.
-Alternatively, any other form of pre-specifying a combination function for the stage-wise p values might be used to control the type-I-error rate.
+In practice, a simple plug-in approach similar to the approach in blinded sample size reassessment might be viable but has not yet been investigated more thoroughly.
+Alternatively, any other form of pre-specifying a combination function for the stage-wise $p$-values might be used to control the type~I-error rate.
 This combination function should then also be optimized over during the planning stage
@@
 
+\subsection*{Acknowledgments}
+
+We would like to thank the editor and an anonymous reviewer for their excellent comments which greatly improved the manuscript.
+
+\subsection*{Data Availability Statement}
+
+The source code required to reproduce all results presented in this manuscript is available at
+\href{https://github.com/kkmann/unblinded-sample-size-adaptation}{github.com}
+and
+\href{https://doi.org/10.5281/zenodo.3925752}{zenodo.org}.
+
+\subsection*{Funding}
+
+This research was supported by the NIHR Cambridge Biomedical Research Centre (BRC-1215-20014).
+The views expressed are those of the authors and not necessarily those of the NHS, the NIHR or the Department of Health and Social Care.
+David S.\ Robertson was funded by the Biometrika Trust and the Medical Research Council under Grant MC\_UU\_00002/6.
+
 \bibliography{manuscript.bib}
 
-\section*{Appendix}
-
-The conditional error for an earlier-than-planned interim analysis ($m'<m$)
-is
-\begin{align}
-    & \Pr_\theta\big[\,Z_{n(Z_m)} > c(Z_m) \cond Z_{m'} = z_{m'}\,\big] \\
-    =& \int \CP_\theta\big(m,n(z_m),z_m, c(z_m), \theta\big)\  p_\theta(z_m \cond z_{m'}) \operatorname{d} z_m
-\end{align}
-where $p_\theta(z_m \cond z_{m'})$ denotes the probability density function of the outcome at $m$ given the partial observed outcome up to $m'$ (see equation~\eqref{eq:joint-distribution}).
-
 \end{document}

### 2021-08-13 R/01-conditional-power-and-friends.R

@@ -3,14 +3,14 @@
 
 
 # Setup ------------------------------------------------------------------------
-suppressPackageStartupMessages({
-    library(tidyverse)
-    library(patchwork)
-})
+options(tidyverse.quiet = TRUE)
+library(tidyverse)
+library(patchwork)
 
 set.seed(42)
-source("util.R")
-dir.create("../output/figures", recursive = TRUE, showWarnings = FALSE)
+
+source("R/util.R")
+dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)
 
 
 
@@ -40,7 +40,7 @@ tibble(
           legend.position = 'top',
           panel.grid.minor = element_blank()
       )
-ggsave("../output/figures/prior_density.pdf", width = 8, height = 3.5)
+ggsave("output/figures/prior_density.pdf", width = 8, height = 3.5)
 
 
 
@@ -74,13 +74,14 @@ tibble(
         aes(z, value, linetype = quantity) +
         geom_line() +
         labs(x = expression(z[m]), y = '') +
-        scale_y_continuous('', breaks = seq(0, 1, by = 0.2)) +
+        scale_y_continuous("conditional power estimate", breaks = seq(0, 1, by = 0.2)) +
         scale_linetype('') +
+    #ylim(c(0, .5)) +
         theme_bw() +
         theme(
             legend.position = 'top'
         )
-ggsave("../output/figures/acp_ocp_pp_example.pdf", width = 8, height = 3.5)
+ggsave("output/figures/acp_ocp_pp_example.pdf", width = 8, height = 3.5)
 
 
 
@@ -107,7 +108,6 @@ expand_grid(
     pivot_longer(c(bias, MAE, MSE), names_to = "quantity") %>%
     ggplot() +
         aes(theta, value, linetype = name) +
-        geom_hline(yintercept = 0) +
         geom_line() +
         facet_wrap(~quantity) +
         labs(x = expression(theta), y = '', color = '') +
@@ -117,7 +117,7 @@ expand_grid(
              legend.position = 'top',
             panel.grid.minor = element_blank()
         )
-ggsave("../output/figures/acp_ocp_pp_bias_mae_mse.pdf", width = 8, height = 3.5)
+ggsave("output/figures/acp_ocp_pp_bias_mae_mse.pdf", width = 8, height = 3.5)
 
 
 
@@ -130,7 +130,8 @@ expand_grid(
            zm = zm + sqrt(m)*theta,
[…]

### 2021-08-13 R/02-naive-recalculation-and-optimal-design.R

@@ -4,18 +4,17 @@
 
 
 # Setup ------------------------------------------------------------------------
-suppressPackageStartupMessages({
-    library(tidyverse)
-    library(patchwork)
-    library(future)
-    library(adoptr)
-})
+options(tidyverse.quiet = TRUE)
+library(tidyverse)
+library(patchwork)
+library(future)
+library(adoptr)
 
 plan(multisession)
 
 set.seed(42)
-source("util.R")
-dir.create("../output/figures", recursive = TRUE, showWarnings = FALSE)
+source("R/util.R")
+dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)
 
 
 
@@ -285,7 +284,7 @@ plt3 <- ggplot(tbl_designs) +
     )
 # compose plot
 plt1 + plt2 + plt3 + plot_layout(guides = "collect") & theme(legend.position = 'top')
-ggsave("../output/figures/naive_adaptive_and_optimal_designs.pdf", width = 8, height = 3.5)
+ggsave("output/figures/naive_adaptive_and_optimal_designs.pdf", width = 8, height = 3.5)
 
 
 

### 2021-08-13 .binder/install.R
switch to conda for binder

@@ -1,9 +1,0 @@
-install.packages("tidyverse")
-install.packages("tinytex")
-install.packages("patchwork")
-install.packages("truncnorm")
-install.packages("adoptr")
-install.packages("nloptr")
-install.packages("rootSolve")
-install.packages("future")
-install.packages("renv")

### 2021-08-14 .binder/install.R
...

@@ -1,0 +1,9 @@
+install.packages("tidyverse")
+install.packages("tinytex")
+install.packages("patchwork")
+install.packages("truncnorm")
+install.packages("adoptr")
+install.packages("nloptr")
+install.packages("rootSolve")
+install.packages("future")
+install.packages("renv")
